---
title: "Advanced forecast evaluation"
order: 9
bibliography: ../nfidd.bib
---

# Introduction

In the last few sessions we built forecasts for many locations at once and scored them with the CRPS.
We scored each location on its own and then averaged.
In this session we look at two things that simple approach misses.
First, scoring locations one at a time is blind to how they move *together*, which the energy score can capture.
Second, a single averaged score, or a significance test, is a fragile way to compare models, and modelling the scores directly is often better.

## Slides

-   [Advanced forecast evaluation](slides/advanced-forecast-evaluation.qmd)

## Objectives

The aim of this session is to score joint forecasts across locations and to treat forecast scores as data worth modelling.

::: {.callout-note collapse="true"}
# Setup

## Source file

The source file of this session is located at `sessions/advanced-forecast-evaluation.qmd`.

## Libraries used

In this session we will use the `fable` package for forecasting, the `dplyr`, `tidyr` and `tsibble` packages for data wrangling, the `ggplot2` package for plotting, the `scoringRules` package for scoring, and the `mgcv` package for modelling the scores.

```{r libraries, message = FALSE}
library("nfidd.forecasting")
library("dplyr")
library("tsibble")
library("fable")
library("tidyr")
library("ggplot2")
library("scoringRules")
library("mgcv")
theme_set(theme_bw())
```

::: callout-tip
The best way to interact with the material is via the [Visual Editor](https://docs.posit.co/ide/user/ide/guide/documents/visual-editor.html) of RStudio.
:::

## Initialisation

We set a random seed for reproducibility.

```{r}
set.seed(406) # for Ted Williams
```

We reuse the same fourth-root transformation and multi-location flu data (`flu_data_hhs`) from the [local hub playground session](hub-playground.qmd).

```{r transform}
fourth_root <- function(x) x^0.25
inv_fourth_root <- function(x) x^4
my_fourth_root <- new_transformation(fourth_root, inv_fourth_root)

data(flu_data_hhs)
```
:::

# From marginal to joint forecasts

A forecast for several regions is not a set of separate numbers.
It is a distribution over the whole *vector* of outcomes, one value per region.
When we score each region on its own and average, we only ever use the **marginal** forecasts.
Two forecasts can share every marginal and still disagree completely about whether regions rise *together*.

That difference matters whenever a decision couples the regions.
If you hold a fixed stockpile to split across regions, a peak that hits everywhere at once is a very different problem from peaks that arrive one region at a time.
Marginal scores cannot see this, because the coupling lives in the joint distribution.

Let's build sample-path forecasts for three regions and pull out the samples we need to score them jointly.

```{r partA-samples}
## fable fits each location separately, so its joint forecast assumes
## independence across regions. A genuinely joint model (e.g. fable::VAR())
## would instead learn the cross-region dependence.
sel_locations <- c("HHS Region 1", "HHS Region 2", "HHS Region 3")
train_cutoff  <- as.Date("2019-02-16")
horizon       <- 4L

fit_joint <- flu_data_hhs |>
  filter(location %in% sel_locations, origin_date <= train_cutoff) |>
  model(arima210 = ARIMA(my_fourth_root(wili) ~ pdq(2, 1, 0)))

sims <- fit_joint |>
  generate(h = 4, times = 1000, bootstrap = TRUE) |>
  group_by(location, .rep) |>
  mutate(horizon = row_number()) |>
  ungroup() |>
  as_tibble()
```

We reshape the samples for our target horizon into a matrix with one row per region and one column per sample, which is the format the scoring functions expect.

```{r partA-matrix}
samp_wide <- sims |>
  filter(horizon == !!horizon) |>
  select(location, .rep, .sim) |>
  pivot_wider(names_from = .rep, values_from = .sim)

target_end_date <- sims |>
  filter(horizon == !!horizon) |> pull(origin_date) |> unique()

dat_indep <- as.matrix(samp_wide[, -1])
rownames(dat_indep) <- samp_wide$location

## observed value at the target date, one per region
y <- flu_data_hhs |> as_tibble() |>
  filter(location %in% sel_locations, origin_date == target_end_date) |>
  arrange(match(location, samp_wide$location)) |>
  pull(wili)
names(y) <- samp_wide$location

target_end_date
y
```

# The energy score

The energy score is a proper scoring rule for a **vector** outcome $\mathbf{y}$, and it generalises the CRPS.
Writing $\mathbf{X}, \mathbf{X}'$ for independent draws from the forecast and $\lVert\cdot\rVert$ for Euclidean distance, it balances accuracy against sharpness exactly as the CRPS does, but reads the whole vector at once.
In one dimension it *is* the CRPS.

::: {.callout-tip .optional collapse="true"}
## Mathematical definition (optional)

$$
\text{ES}(F, \mathbf{y}) = \mathbb{E}\lVert \mathbf{X} - \mathbf{y} \rVert - \tfrac{1}{2}\,\mathbb{E}\lVert \mathbf{X} - \mathbf{X}' \rVert
$$

The first term rewards forecasts whose draws land close to the observed vector.
The second term rewards a concentrated cloud of draws.
To compute it we replace the expectations with averages over our predictive samples.
:::

::: callout-tip
### Key things to note about the energy score

-   Small values are better, and it is reported in the units of the data.
-   It is a genuine generalisation of the CRPS, to which it reduces in one dimension.
-   It needs **joint samples**, so a forecast that treats regions independently and one that couples them are scored differently.
-   Like the CRPS it is best used to **compare** forecasts, not read in isolation.
:::

::: {.callout-tip .optional collapse="true"}
## Interactive: the energy score

```{=html}
<iframe class="nfidd-embed" src="../interactive/advanced-forecast-evaluation.html?embed=energy" title="Energy score interactive" loading="lazy" style="width:100%;height:820px;border:1px solid #e5e7eb;border-radius:10px;background:#fff"></iframe>
<script>(function(){if(window.__nfiddEmbedInit)return;window.__nfiddEmbedInit=1;window.addEventListener('message',function(ev){var h=ev&&ev.data&&ev.data.nfiddEmbedHeight;if(!h)return;document.querySelectorAll('iframe.nfidd-embed').forEach(function(f){if(f.contentWindow===ev.source)f.style.height=(h+24)+'px';});});})();</script>
```

Drag the observation and rotate the forecast cloud with the dependence slider.
The mean marginal CRPS never moves; only the energy score responds to the dependence.
:::

To see the difference dependence makes, we build a second forecast with **identical marginals** but perfect positive dependence.
Sorting each region's samples makes high draws in one region line up with high draws in the others.

```{r partA-dependence}
## identical marginals, but now the regions move together (comonotonic)
dat_comono <- t(apply(dat_indep, 1, sort))

## marginal CRPS (per region) is unchanged by the reordering
crps_indep  <- scoringRules::crps_sample(y = y, dat = dat_indep)
crps_comono <- scoringRules::crps_sample(y = y, dat = dat_comono)

rbind(independent = crps_indep, comonotonic = crps_comono)
c(mean_indep = mean(crps_indep), mean_comono = mean(crps_comono))
```

The per-region CRPS, and therefore the average, are exactly the same for both forecasts.
The energy score is not.

```{r partA-energy}
es_indep  <- scoringRules::es_sample(y = y, dat = dat_indep)
es_comono <- scoringRules::es_sample(y = y, dat = dat_comono)

c(independent = es_indep, comonotonic = es_comono)
```

::: callout-tip
## Take 5 minutes

The mean marginal CRPS is identical for the two forecasts, but the energy score prefers the comonotonic one.
Look at the observed vector `y` relative to the forecast medians.
Why does coupling the regions help here, and when would it hurt instead?
:::

::: {.callout-note collapse="true"}
## Solution

At this target date all three regions came in **below** their forecast medians at the same time, a synchronised late-season decline.
The comonotonic forecast puts probability on regions being jointly low (or jointly high), so it places mass near what actually happened and scores better.
The independent forecast, which is what fitting each region separately gives us, spreads its joint mass across combinations where one region is high while another is low, and scores worse for it.

If instead the regions had diverged, one high while another was low, the coupling would have hurt and the independent forecast would have scored better.
The marginal CRPS is blind to all of this: it is identical either way.
The lesson is not that dependence is always good, but that the energy score can see it and marginal scores cannot.
:::

# Modelling scores

When we compare models we usually reach for a single averaged score, and sometimes a significance test on top.
Both are fragile.

::: callout-tip
## Take 2 minutes

Before reading on, write down two reasons a single average CRPS could give a misleading comparison of models.
:::

::: {.callout-note collapse="true"}
## Solution

-   Scores are strongly right-skewed, so a few hard weeks dominate the mean.
-   They vary with the scale of the data, so big locations dominate small ones.
-   They are correlated across dates, locations and horizons, so there is much less information than the number of rows suggests.
-   Forecasts differ in difficulty, so an average rewards whoever predicted the easy targets.
-   A significance test tells you *whether* two models differ, not *where* or *by how much*.
:::

A more useful approach is to treat the scores themselves as data and model them.
We use a set of CRPS scores that we have pre-computed from a small backtest of three contrasting models over several forecast dates.

```{r partB-data}
data(advanced_scores)
advanced_scores
```

::: {.callout-note collapse="true"}
## How these scores were produced

The scores come from a time-series cross-validation backtest of three models (a random walk, an ARIMA(2,1,0), and an ARIMA(2,0,0) plus Fourier terms) for three HHS regions over eight forecast dates in the 2018/2019 season, scored with the CRPS.
This follows the same [time-series cross-validation](forecast-evaluation.qmd#time-series-cross-validation) and `scoringutils` scoring you saw in the earlier sessions.
The full code is in `data-raw/advanced_scores.R`.
:::

Here is the naive summary: the average CRPS per model.

```{r partB-naive}
advanced_scores |>
  group_by(model) |>
  summarise(mean_crps = mean(crps), sd_crps = sd(crps), n = n()) |>
  arrange(mean_crps)

## the scores are strongly right-skewed
quantile(advanced_scores$crps, c(.1, .5, .9, .99))
```

The `fourier` model has the lowest average, but the standard deviations are large and the scores are heavily skewed.
Now we model the scores instead.
We adjust for horizon and location difficulty, and we get uncertainty on each model's effect.

::: callout-important
## Do not transform the score itself

It is tempting to fit `lm(log(crps) ~ ...)` to handle the skew, but this breaks what made the CRPS worth using.
The CRPS is a *proper* score, so the average CRPS is minimised by the better forecaster.
The log is nonlinear, so the average *log* CRPS is not, and ranking models by it is no longer a proper comparison.

The fix is to keep the CRPS on its natural scale as the response, and use a **generalised additive model** with a Gamma family and a log **link**.
The link handles positivity and skew, but the model still describes the mean CRPS, so the comparison stays proper.
If instead you want to reward performance on the log scale, transform the *forecasts and observations* before scoring (the `transform = log1p` route from the [real-world forecasts session](real-world-forecasts.qmd#log-scale-scores)), which is still proper, and then model those scores.[@bosse2023scorin]
:::

```{r partB-model}
sdf <- advanced_scores |>
  mutate(model = relevel(factor(model), ref = "arima210"),
         location = factor(location))

## Gamma family + log link models E[CRPS] on its natural scale (the response is
## never transformed), so the model comparison stays proper. The smooth lets
## the score vary flexibly with horizon.
score_model <- mgcv::gam(
  crps ~ model + location + s(horizon, k = 4),
  family = Gamma(link = "log"), data = sdf
)
round(summary(score_model)$p.table, 3)
```

The `model` terms are what we care about.
The smooth term for horizon confirms that horizon drives most of the variation in scores.

```{r partB-smooth}
round(summary(score_model)$s.table, 3)
```

We can turn the model coefficients into an adjusted mean CRPS per model with an interval, holding horizon and location fixed.

```{r partB-adjusted}
newd <- data.frame(
  model    = factor(levels(sdf$model), levels = levels(sdf$model)),
  location = factor("HHS Region 1", levels = levels(sdf$location)),
  horizon  = 2
)
pr <- predict(score_model, newd, se.fit = TRUE)  # link (log) scale
adj_means <- data.frame(
  model    = as.character(newd$model),
  adj_crps = exp(pr$fit),
  lower    = exp(pr$fit - 1.96 * pr$se.fit),
  upper    = exp(pr$fit + 1.96 * pr$se.fit)
)
adj_means
```

A forest plot makes the overlap easy to read.

```{r partB-forest}
adj_means |>
  mutate(model = reorder(model, adj_crps)) |>
  ggplot(aes(x = adj_crps, y = model)) +
  geom_pointrange(aes(xmin = lower, xmax = upper)) +
  labs(x = "Adjusted mean CRPS (horizon 2, Region 1)", y = NULL)
```

::: callout-tip
## Take 5 minutes

Compare the naive ranking with the model coefficients and the adjusted intervals.
Does the model that "wins" on average clearly beat the others?
What does the model tell you that the average does not?
:::

::: {.callout-note collapse="true"}
## Solution

The naive ranking puts `fourier` first, but the score model shows its edge is not distinguishable from noise, with a `modelfourier` coefficient that is small relative to its standard error.
The `rw` model is effectively tied with `arima210`.
The adjusted intervals for all three models overlap heavily, so the average overstates a difference that is within noise.

What the model adds is structure.
The smooth term shows horizon dominates everything, with four-week-ahead scores far worse than one-week-ahead, and Region 2 is systematically harder to predict.
These are the effects that a single average blends together and hides.
:::

::: callout-tip
The standard errors above are still optimistic, because the score rows are correlated: they share origin dates and locations, and horizons are nested within each forecast.
Adding random effects for date, location and horizon is one way to handle this, either as a mixed-effects GAM or a Bayesian model, and is the approach taken in large multi-model comparisons.[@sherrattPredictivePerformanceMultimodel2023;@lopezChallengesCOVID19Case2024]
However, even these models may fail to account for all sources of correlation, leading to misspecified standard errors in your model.
:::

# Going further

## Challenge

-   Recompute the energy-score comparison at a target date where the regions diverged rather than declining together. Does the independent forecast now score better?
-   Fit a genuinely joint model with `fable::VAR()` and compare its energy score to the independent ARIMA forecast.
-   Add the variogram score (`scoringRules::vs_sample()`) to the joint comparison. How does it differ from the energy score in what it rewards?
-   Let the horizon smooth vary by model in the score model (`s(horizon, by = model)`). Do any models do relatively better at longer horizons?

## Methods in practice

-   The `scoringRules` package [documentation](https://cran.r-project.org/package=scoringRules) covers multivariate scores including the energy and variogram scores.
-   Hubs increasingly collect sample (trajectory) forecasts rather than only quantiles, which keeps the joint distribution across locations and horizons.[@hubverseModelOutput2026]

# Wrap up

-   Review what you've learned in this session with the [learning objectives](../reference/learning_objectives.qmd)

::: {#refs}
:::
