---
title: "Baselines and seasonality"
order: 8
bibliography: ../nfidd.bib
---

```{r echo=FALSE}
options(width=100)
```

# Introduction

In the [previous session](time-series-101) we built some simple ARIMA models and used a fourth-root transformation to keep our forecasts positive.
Before we try to build *better* models, though, we need something to compare them against.
In this session we introduce **baseline** (reference) models — deliberately simple, naive forecasts — and then treat **seasonality** as one more reference point before combining it with recent dynamics to (hopefully) improve on all of them.

## Slides

-   [Baselines and seasonality](slides/baselines-and-seasonality)

## Objectives

The goal of this session is to introduce baseline/reference models, to build a seasonal baseline for the ILI data, to see whether adding recent dynamics improves on the baselines, and to use residual diagnostics to judge whether a model has captured the structure in the data.

::: callout-caution
As with the last session, none of the models introduced in this section are designed for real-world use!
:::

::: {.callout-note collapse="true"}
# Setup

## Source file

The source file of this session is located at `sessions/baselines-and-seasonality.qmd`.

## Libraries used

In this session we will use the `nfidd.forecasting` package for some flu data, the `fable` package to fit and forecast models, the `dplyr` package for data wrangling, and the `ggplot2` and `ggtime` packages for plotting.

```{r libraries, message = FALSE}
library("nfidd.forecasting")
library("fable")
library("dplyr")
library("ggplot2")
library("ggtime")
```

::: callout-tip
The best way to interact with the material is via the [Visual Editor](https://docs.posit.co/ide/user/ide/guide/documents/visual-editor.html) of RStudio.
:::

## Initialisation

We set a random seed for reproducibility, load the same flu data as in the previous session, and re-create the fourth-root transformation we defined there.

```{r}
set.seed(123)
data(flu_data)

fourth_root <- function(x) x^0.25
inv_fourth_root <- function(x) x^4
my_fourth_root <- new_transformation(fourth_root, inv_fourth_root)
```
:::

# Baseline (reference) models

Before we ask "is my model any good?", we have to ask "good *compared to what*?".
A **baseline** (or **reference**) model is a deliberately simple, naive forecast that any genuinely useful model ought to beat.
Baselines are worth taking seriously because:

-   they set a floor — if a complicated model can't beat a one-line baseline, something is wrong;
-   they are transparent, fast, and hard to overfit;
-   in collaborative forecasting (later in the course) a *shared* baseline lets everyone measure their "skill" against the same yardstick.

Three simple ideas come up again and again:

-   **Persistence** ("nothing changes"): forecast every future week as the *most recent observation*.
-   **Mean** ("it reverts to normal"): forecast every future week as the *long-run average* of the data.
-   **Seasonal** ("this time of year looks like it usually does"): forecast the *typical seasonal pattern*, ignoring recent wiggles.

We'll build all three, keeping the fourth-root transformation from the last session so that every model is on the same footing (and none of them can predict a negative percentage).

## Persistence and mean baselines

The two simplest baselines don't need any notion of season at all:

-   **Persistence** is `NAIVE()`: carry the last observed value forward.
-   **Mean** is `MEAN()`: forecast the average of all past values.

```{r fit-simple-baselines}
simple_baselines <- flu_data |>
  model(
    persistence = NAIVE(my_fourth_root(wili)),
    mean        = MEAN(my_fourth_root(wili))
  )

forecast(simple_baselines, h = 40) |>
  autoplot(flu_data |> filter(epiweek >= as.Date("2015-09-01"))) +
  facet_wrap(~.model) +
  labs(title = "Two simple baselines", y = "% of visits")
```

Neither is a *good* forecast of a flu season — persistence just extends the last value in a flat line, and the mean ignores where we currently are entirely — but they are exactly the kind of honest reference points we want to beat.

::: callout-note
A **random walk** (`RW()`) produces the *same point forecast* as persistence — the last observed value — and differs only in how quickly its prediction intervals fan out over the forecast horizon.
Here we use `NAIVE()` for the persistence point forecast and `MEAN()` for the "reverts to the average" baseline; you'll meet the random walk again as a baseline in the [forecast evaluation session](forecast-evaluation).
:::

## A seasonal baseline

Our third baseline is smarter about *where in the year* we are, but is still simple: it captures only the seasonality, and nothing about recent dynamics.

There is a whole part of ARIMA modeling that accounts for seasonality (these are called Seasonal ARIMA, or SARIMA, models).
However, using seasonal ARIMA models with weekly data is complicated, because some years have 52 weeks and some have 53.
Here are two relatively straight-forward ways you could address this:

1.  Define a "season-year" where "week 1" is the start of each season (say, the first week of September) and count to 52 from there, dropping any 53rd week. Model with a periodicity of 52.
2.  Include [Fourier series terms](https://en.wikipedia.org/wiki/Fourier_series#Synthesis) (sums of sine and cosine terms with a fixed periodicity of 365 days) to fit annual periodic trends.

We're going to use approach 2 because it's a bit easier to get up and running in `fable`: you can just add a `fourier()` term to the model equation and `fable` will add the sine and cosine predictors for you.
You can read more about this approach in [FPP3 Chapter 12.1](https://otexts.com/fpp3/complexseasonality.html#dynamic-harmonic-regression-with-multiple-seasonal-periods).

A model with *only* Fourier terms and no ARIMA dynamics (`pdq(0, 0, 0)`) is a **harmonic regression**: a regression of the (transformed) data on sine and cosine curves with an annual period.

$$ g(y_t) = \sum_{k=1}^{3} \left[ \alpha_k \cos\left( \frac{2\pi \cdot k \cdot d(t)}{365} \right) + \beta_k \sin\left( \frac{2\pi \cdot k \cdot d(t)}{365} \right) \right] + \varepsilon_t $$

where $g(\cdot)$ is the fourth-root transformation, $d(t)$ returns the day of the year of the first day of week $t$, $\alpha_k$ and $\beta_k$ are the Fourier coefficients that `fable` estimates, and the number of harmonics $K = 3$ is chosen (somewhat arbitrarily) to allow a flexible but smooth annual shape.

```{r fit-fourier}
fit_fourier <- flu_data |>
  model(fourier = ARIMA(my_fourth_root(wili) ~ pdq(0, 0, 0) + fourier(period = "year", K = 3)))
report(fit_fourier)

forecast(fit_fourier, h = 40) |>
  autoplot(flu_data |> filter(epiweek >= as.Date("2015-09-01"))) +
  labs(title = "Seasonal baseline (harmonic regression)",
       y = "% of visits")
```

Because it has no auto-regressive, differencing, or moving-average terms, this model's forecast is driven *entirely* by the estimated seasonal curve: it is essentially a smooth "seasonal climatology" that repeats the average shape of a season.
Like the other baselines, it pays no attention to the most recent observations — its forecast for next week would be the same whether the last few weeks were unusually high or low.

::: callout-tip
We will re-use this exact `fourier` model later in the course as a simple **seasonal baseline** to compare more sophisticated models against.
:::

# Improving on the baselines: adding recent dynamics

Each baseline uses just one idea. Persistence and the mean react only to the level of the data (and not even the recent level, in the case of the mean); the seasonal baseline knows the shape of a typical season but ignores where we are right now.
Can we do better than *all three* by combining seasonality with recent dynamics?
A natural step is to add the Fourier terms to an AR(2) model, so the forecast is anchored to recent values *and* pulled toward the seasonal pattern:

$$ g(y_t) = \phi_1 g(y_{t-1}) + \phi_2 g(y_{t-2}) + \sum_{k=1}^{3} \left[ \alpha_k \cos\left( \frac{2\pi \cdot k \cdot d(t)}{365} \right) + \beta_k \sin\left( \frac{2\pi \cdot k \cdot d(t)}{365} \right) \right] + \varepsilon_t $$

# Comparing the forecasts

Let's put the three baselines and our combined model side by side:

```{r fit-all}
fits <- flu_data |>
  model(
    persistence      = NAIVE(my_fourth_root(wili)),
    mean             = MEAN(my_fourth_root(wili)),
    fourier          = ARIMA(my_fourth_root(wili) ~ pdq(0, 0, 0) + fourier(period = "year", K = 3)),
    arima200_fourier = ARIMA(my_fourth_root(wili) ~ pdq(2, 0, 0) + fourier(period = "year", K = 3))
  )

forecast(fits, h = 40) |>
  autoplot(flu_data |> filter(epiweek >= as.Date("2015-09-01"))) +
  facet_wrap(~.model) +
  labs(title = "Three baselines and an improved model", y = "% of visits")
```

The combined `arima200_fourier` model captures the seasonal shape *and* starts from roughly the right recent level — visually an improvement over all three baselines.
But "looks better" is not good enough: in the next session we'll learn to *measure* forecast performance with proper scoring rules, and put the "did we actually beat the baselines?" question on a firm footing.

::: callout-tip
## Take 5 minutes

Look at the four panels. Which baseline do you think will be hardest to beat, and *when* during the season? Where does the combined model most clearly improve on each baseline?
:::

# Checking the residuals

A visual sanity check of the forecast is a good start, but a more rigorous question is: *has the model captured all of the structure in the data?*
If it has, then what's left over — the model **residuals** (technically the one-step-ahead "innovation" residuals) — should look like white noise: no trend, roughly constant variance, and no left-over autocorrelation.
Any pattern remaining in the residuals is signal the model failed to use.

`ggtime::gg_tsresiduals()` gives us the three standard diagnostic views at once. Let's look at our combined model:

```{r residuals}
fits |>
  select(arima200_fourier) |>
  gg_tsresiduals()
```

::: callout-tip
## Take 5 minutes

Look at the three panels. Do the residuals look like white noise? In particular, is there any left-over autocorrelation in the ACF panel — especially around lag 52 (one year)?
:::

::: {.callout-note collapse="true"}
## Solution

-   The **top panel** (residuals over time) should hover around zero, but you'll likely still see that the variance is larger during the winter peaks — a sign that our constant-variance assumption isn't perfect even after the fourth-root transform.
-   The **ACF panel** (bottom-left) is the key one: some bars still poke outside the dashed significance bounds (at short lags and near the seasonal lag), which tells us the model has *not* captured all the autocorrelation — there is still structure that a better model could exploit.
-   The **histogram** (bottom-right) should be roughly bell-shaped and centred on zero.

So the model is a clear improvement on the baselines, but the residual diagnostics show it is still not "done" — exactly the kind of clue that guides the next round of model building.
:::

# End-of-session reflection: Bayesian vs. Frequentist modeling

The statistical models we've discussed in this session represent just a subset of available forecasting approaches.
These models offer a different workflow for thinking about modeling than generative models. There are some good examples of generative models in [the companion nowcasting course](https://nfidd.github.io/sismid-nowcasting/) that we taught earlier this week.
Whilst Bayesian approaches work forward from domain knowledge to specify data-generating processes, traditional time series methods such as these ARIMA models work backward from observed patterns using standard toolkits (transformations, differencing, residual analysis) to achieve stationarity and improve predictions.
Both approaches ultimately aim to capture the same underlying data-generating process but from different perspectives.
Something to think about!

# Going further

-   Which of the baselines is hardest to beat, and does that change over the course of a season? (We'll answer this quantitatively in the evaluation session.)
-   The other approach to seasonality mentioned above (a "season-year" re-indexing, i.e. a genuine seasonal ARIMA / SARIMA model) — try implementing it and compare it to the Fourier approach.
-   We fixed $K = 3$ Fourier harmonics somewhat arbitrarily. Try $K = 1, 2, 4$ and compare the fits (e.g. with `glance()` and the reported information criteria). What happens to the seasonal shape as $K$ grows?
-   Read about [dynamic harmonic regression](https://otexts.com/fpp3/dhr.html) and [residual diagnostics](https://otexts.com/fpp3/diagnostics.html) in the FPP3 textbook.

# Wrap up

-   Review what you've learned in this session with the [learning objectives](../reference/learning_objectives)
-   Share your [questions and thoughts](../reference/help)

## References

::: {#refs}
:::
