---
title: "Epidemiologically-motivated forecasting"
order: 5
bibliography: ../nfidd.bib
---

```{r echo=FALSE}
options(width=100)
```

# Introduction

In earlier sessions we built forecasts from purely statistical models such as ARIMA.
These models find patterns in reported counts and project those patterns forward.
They do not know anything about *how* infections arise, how they are transmitted, or why an epidemic must eventually decline.

In this session we add epidemiological mechanism to the forecast.
The key idea is the **renewal equation**, a model of incidence that is mathematically an autoregression whose coefficients are the generation-time distribution and whose scale is the reproduction number $R_t$ [@bhattSemiMechanisticBayesianModeling2020; @coriNewFrameworkSoftware2013].
This gives the forecast access to information that a pure time-series model cannot use.

We will account for the delay between infection and case report, and for susceptible depletion that bends the epidemic curve over at the peak.
We then backtest a set of models with different amounts of epidemiological mechanism, making forecasts from a series of past dates and scoring them against what happened next.
This asks a question about parsimony and identification.
Does the model that matches the data-generating process win, and do unnecessary parameters hurt?
Finally we deliberately break the mechanism with a wrong population size to see the cost of a wrong model up close.

::: callout-note
## EpiNow2 is the tool, not the point
We fit these models with the EpiNow2 R package [@abbottEpiNow2EstimateRealtime2025].
EpiNow2 combines the renewal equation, the delays from infection to report, and optional susceptible depletion, and fits them in Stan.
The same model structure fit with other software would give the same forecasts, so the focus here is on the structure rather than the package.
:::

## Slides

-   [Epidemiologically-motivated forecasting](slides/epi-motivated-forecasting.qmd)

## Objectives

The aim of this session is to introduce epidemiologically-motivated forecasting using renewal equations, to connect the renewal model to the autoregressive models from earlier sessions, and to backtest models with different levels of epidemiological mechanism against each other.

::: {.callout-note collapse="true"}
# Setup

## Source file

The source file of this session is located at `sessions/epi-motivated-forecasting.qmd`.

## Libraries used

In this session we will use the `EpiNow2` package for renewal-equation forecasting, the `fable` package for the ARIMA baseline, the `dplyr`, `tidyr`, `tibble`, and `purrr` packages for data wrangling, the `ggplot2` and `gridExtra` packages for plotting, and the `scoringutils` package for forecast evaluation.

```{r libraries, message = FALSE}
library("nfidd.forecasting")
library("EpiNow2")
library("fable")
library("dplyr")
library("tidyr")
library("tibble")
library("purrr")
library("ggplot2")
library("gridExtra")
library("scoringutils")
theme_set(theme_bw())
```

::: callout-tip
The best way to interact with the material is via the [Visual Editor](https://docs.posit.co/ide/user/ide/guide/documents/visual-editor.html) of RStudio.
:::

## Initialisation

We set a random seed for reproducibility.
Setting this ensures that you should get exactly the same results on your computer as we do.
This is not strictly necessary but will help us talk about the models.

```{r}
set.seed(7321)
```

We also define the fourth-root transformation used in earlier sessions, so that our ARIMA baseline is the same model family students already know.

```{r}
fourth_root <- function(x) x^0.25
inv_fourth_root <- function(x) x^4
my_fourth_root <- new_transformation(fourth_root, inv_fourth_root)
```

We also set the inference configuration that every EpiNow2 fit in this session will use, so we can change it in one place if we want different inference later.

```{r}
stan_config <- stan_opts(method = "vb", backend = "rstan", samples = 1000)
```
:::

::: {.callout-note collapse="true"}
## Inference in this session, and how to switch to pathfinder or MCMC

EpiNow2 fits a Bayesian model in Stan, and we refit it dozens of times across forecast dates and models.
Full Markov-chain Monte Carlo (MCMC) sampling gives the most honest posterior but takes around 40 seconds per fit, which would make the backtest take 20 minutes or more.
Variational Bayes (`vb`) gives an approximate posterior in one to two seconds via the `rstan` backend that ships with EpiNow2, so you do not need to install anything extra.
This is the right trade-off for a teaching backtest where we care about relative model behaviour rather than calibrated uncertainty.

If you have the `cmdstanr` package and a CmdStan install, pathfinder usually gives a better approximation than `vb`.
Switch by changing the single `stan_config` line above:

```r
stan_config <- stan_opts(method = "pathfinder", backend = "cmdstanr")
```

For full MCMC (best-calibrated, slowest):

```r
stan_config <- stan_opts(samples = 1000, chains = 2, warmup = 300, method = "sampling", backend = "rstan")
```

In a real-world analysis you would use more warmup and sampling iterations for better-calibrated uncertainty.
We use reduced settings here to keep the session runtime manageable.
Every fit in the session picks up the change automatically.
:::

# From autoregression to the renewal equation

An ARIMA model predicts the next value of a time series as a weighted sum of its past values.
A renewal equation has the same autoregressive shape, but every piece is given epidemiological meaning.

:::: {.columns}
::: {.column width="48%"}
#### ARIMA

$$ y_t = \sum_s \phi_s\, y_{t-s} + \varepsilon_t $$

- Coefficients $\phi_s$ are free, estimated from the data
- $\varepsilon_t$ is an estimated noise term
- No epidemiological meaning
:::

::: {.column width="48%"}
#### Renewal

$$ I_t \sim \mathrm{Poisson}\!\left(R_t \sum_s w_s\, I_{t-s} + \iota_t\right) $$

- Coefficients are $R_t\, w_s$, the reproduction number times the generation-time weight
- $\iota_t$ is imported infections, set to zero here
- Every term is an epidemiological quantity
:::
::::

Here $I_t$ is the number of new infections, $R_t$ is the reproduction number, and $w_s$ is the **generation-time distribution**, the probability that an infection caused $s$ days ago transmits now [@wallingaDifferentEpidemicCurves2004].
The generation-time weights $w_s$ are not free but must form a valid probability distribution, and are typically estimated from data on serial intervals or generation times, not from the incidence curve itself.

# Simulating an outbreak

We simulate a single outbreak to forecast, using `simulate_infections()` from EpiNow2.
We set a constant reproduction number above 1 and a fixed population size, which produces an epidemic that grows, peaks, and then declines as susceptibles are depleted.
Setting `pop_period = "all"` applies the depletion across the whole simulated period, so the turnover is built into the data-generating process.
The generation time and delays are the example distributions supplied with EpiNow2, fixed to point values with `fix_parameters()` because simulation requires known parameters.

```{r simulate}
gen_time <- fix_parameters(example_generation_time)
delays <- fix_parameters(example_incubation_period + example_reporting_delay)
pop_size <- 100000
dates <- seq.Date(as.Date("2023-01-01"), by = "day", length.out = 170)
R <- data.frame(date = dates, R = 1.25)

sim <- simulate_infections(
  R = R,
  initial_infections = 20,
  generation_time = gt_opts(gen_time),
  delays = delay_opts(delays),
  obs = obs_opts(family = "poisson"),
  pop = Fixed(pop_size),
  pop_period = "all"
)

reported <- sim |>
  as_tibble() |>
  filter(variable == "reported_cases") |>
  transmute(date, confirm = value)
```

The simulated series of reported cases rises to a peak in late March before declining.

```{r plot-outbreak}
ggplot(reported, aes(x = date, y = confirm)) +
  geom_col(fill = "#4292C6") +
  labs(x = "Date", y = "Reported cases", title = "Simulated outbreak") +
  theme_bw()
```

The decline happens because susceptibles are running out.
The population at risk is finite, so once enough people have been infected the effective reproduction number drops below 1 even though the underlying $R_t$ is still 1.25.
This is the mechanism we want our forecasts to capture.

# A single forecast

To forecast we truncate the series at a forecast date and fit only to the data available up to that point.
We choose a date shortly before the peak, where the epidemic is still growing and the turnover has not yet been observed.

```{r truncate}
forecast_date <- as.Date("2023-03-18")
train <- reported |> filter(date <= forecast_date)
```

We fit two models, the EpiNow2 default renewal model and the ARIMA baseline from earlier sessions.
Both use a 14-day forecast horizon.

```{r single-forecast, warning = FALSE, message = FALSE}
en2_default <- estimate_infections(
  train,
  generation_time = gt_opts(example_generation_time),
  delays = delay_opts(example_incubation_period + example_reporting_delay),
  rt = rt_opts(rw = 7),
  gp = NULL,
  forecast = forecast_opts(horizon = 14),
  stan = stan_config
)

arima_default <- train |>
  as_tsibble(index = date) |>
  model(ar2 = ARIMA(my_fourth_root(confirm)))
```

We plot both forecasts against the data that were later observed.
The shaded ribbons are the 50% and 90% credible intervals, the lines are the posterior medians, and the points are the observations split into the data used for fitting and the held-out truth.

```{r plot-single-forecast}
en2_fc_sum <- summary(en2_default, type = "parameters") |>
  as_tibble() |>
  filter(variable == "reported_cases", type == "forecast") |>
  transmute(date, model = "epi_default", median, lower_50, lower_90, upper_50, upper_90)

arima_fc_sum <- generate(arima_default, h = 14, times = 1000) |>
  as_tibble() |>
  group_by(date) |>
  summarise(
    median = median(.sim, na.rm = TRUE),
    lower_50 = quantile(.sim, 0.25, na.rm = TRUE),
    lower_90 = quantile(.sim, 0.05, na.rm = TRUE),
    upper_50 = quantile(.sim, 0.75, na.rm = TRUE),
    upper_90 = quantile(.sim, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(model = "arima")

single_fc_sum <- bind_rows(en2_fc_sum, arima_fc_sum)

ggplot(single_fc_sum, aes(x = date, fill = model, colour = model)) +
  geom_ribbon(aes(ymin = lower_90, ymax = upper_90), alpha = 0.15) +
  geom_ribbon(aes(ymin = lower_50, ymax = upper_50), alpha = 0.3) +
  geom_line(aes(y = median)) +
  geom_point(
    data = reported |> filter(date > forecast_date - 21, date <= forecast_date + 14),
    aes(x = date, y = confirm),
    inherit.aes = FALSE
  ) +
  geom_vline(xintercept = forecast_date, linetype = "dashed") +
  scale_fill_manual(values = c("epi_default" = "#4292C6", "arima" = "#E6550D")) +
  scale_colour_manual(values = c("epi_default" = "#4292C6", "arima" = "#E6550D")) +
  labs(
    x = "Date", y = "Reported cases",
    title = "14-day forecast from just before the peak"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

Neither model knows that the epidemic is about to turn over.
The ARIMA projects the recent trend forward, and the renewal model does the same because this model has no mechanism for susceptible depletion.
Both over-shoot the peak.
This foreshadows the backtest, where we give the EpiNow2 models the ability to see the turnover coming.

::: callout-tip
## Take 5 minutes

Look at the two forecast ribbons against the held-out observations.
Which model looks closer to the truth over the first few days, and which drifts further off as the horizon grows?
What single piece of information would either model need to predict the turning point?
:::

# The models we will compare

We compare four models that differ in how much epidemiological mechanism and flexibility they include.
This is a story about parsimony and identification.
The data-generating process has constant $R_t$ and a known population, so the model that matches it should win.
Unnecessary flexibility (a random walk on $R_t$) and an unidentified parameter (estimating the population) should hurt.

- **ARIMA**. `ARIMA(my_fourth_root(confirm))` with auto-ARIMA, so fable selects the order automatically from the data. It may select different orders at different forecast dates, which is a feature of a fair statistical baseline rather than a bug.
- **constant_rt_fixed**. Constant reproduction number (the default `rw = 0`) with the *true* population size. This matches the data-generating process of constant $R_t$ plus susceptible depletion, so it should win.
- **rw_fixed**. Adds a weekly random walk on $R_t$ (`rw = 7`) with the true population. The random walk is unnecessary flexibility because the data-generating process has constant $R_t$.
- **rw_estimated**. Weekly random walk on $R_t$ with an estimated population (`Normal(mean = pop_size, sd = pop_size / 2)`). The population is weakly identified from a single outbreak, so estimating it adds uncertainty without benefit.

The EpiNow2 models share the same renewal structure and differ only in the `rt_opts()` arguments.
The ARIMA baseline is fit separately with `fable`, using the same fourth-root transformation as in earlier sessions.

The wrong-population model is deliberately left out of the backtest.
It appears later as a standalone demonstration.

```{r model-specs}
epi_models <- list(
  constant_rt_fixed = rt_opts(pop = Fixed(pop_size), future = "latest", pop_period = "all"),
  rw_fixed          = rt_opts(rw = 7, pop = Fixed(pop_size), future = "latest", pop_period = "all"),
  rw_estimated      = rt_opts(rw = 7, pop = Normal(mean = pop_size, sd = pop_size / 2), future = "latest", pop_period = "all")
)
```

# Backtesting across forecast dates

We backtest each model from 10 forecast dates, one week apart, spanning the growth phase, the peak, and the decline.
Contrasting phases matters because susceptible depletion only shapes the forecast near and after the peak.
Early in the growth phase it is almost irrelevant.

```{r forecast-dates}
forecast_dates <- seq.Date(
  from = as.Date("2023-02-15"), to = as.Date("2023-04-19"), by = "week"
)
length(forecast_dates)
```

We define a helper that fits an EpiNow2 model with variational Bayes and returns sample-level predictions, and another that does the same for the ARIMA baseline via `fable::generate()`.

```{r fit-helpers}
fit_epinow2 <- function(rt, data) {
  estimate_infections(
    data,
    generation_time = gt_opts(example_generation_time),
    delays = delay_opts(example_incubation_period + example_reporting_delay),
    rt = rt, gp = NULL,
    forecast = forecast_opts(horizon = 14),
    stan = stan_config
  )
}

forecast_arima_samples <- function(train, horizon = 14, times = 1000) {
  fit <- train |>
    as_tsibble(index = date) |>
    model(ar2 = ARIMA(my_fourth_root(confirm)))
  generate(fit, h = horizon, times = times) |>
    as_tibble() |>
    transmute(
      forecast_date = max(train$date),
      date,
      horizon = as.integer(date - max(train$date)),
      sample = as.integer(.rep),
      predicted = .sim,
      model = "arima"
    )
}
```

We loop over the forecast dates, fitting all four models at each date and combining the sample-level predictions into a single tibble.

```{r backtest, results = "hide", warning = FALSE, message = FALSE}
backtest_one <- function(fdate) {
  train <- reported |> filter(date <= fdate)
  en2 <- map_dfr(names(epi_models), function(m) {
    fit <- fit_epinow2(epi_models[[m]], train)
    get_predictions(fit, format = "sample") |>
      as_tibble() |>
      transmute(
        forecast_date = fdate,
        date, horizon, sample, predicted, model = m
      )
  })
  arm <- forecast_arima_samples(train)
  bind_rows(en2, arm)
}

all_forecasts <- map_dfr(forecast_dates, backtest_one) |>
  filter(!is.na(predicted))
```

## Visualising the backtested forecasts

Before scoring, we look at the forecast trajectories themselves.
Each panel is a forecast date.
The ribbons are the 50% and 90% credible intervals, the lines are the posterior medians, and the points are the observed data.

```{r plot-backtests, fig.height = 10}
trajectory_summary <- all_forecasts |>
  filter(horizon > 0) |>
  group_by(model, forecast_date, date) |>
  summarise(
    median = median(predicted, na.rm = TRUE),
    lower_50 = quantile(predicted, 0.25, na.rm = TRUE),
    lower_90 = quantile(predicted, 0.05, na.rm = TRUE),
    upper_50 = quantile(predicted, 0.75, na.rm = TRUE),
    upper_90 = quantile(predicted, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

obs_windows <- map_dfr(forecast_dates, function(fd) {
  reported |>
    filter(date >= fd - 28, date <= fd + 14) |>
    mutate(forecast_date = fd)
})

ggplot(trajectory_summary, aes(x = date)) +
  geom_ribbon(aes(ymin = lower_90, ymax = upper_90, fill = model), alpha = 0.15) +
  geom_ribbon(aes(ymin = lower_50, ymax = upper_50, fill = model), alpha = 0.3) +
  geom_line(aes(y = median, colour = model)) +
  geom_point(data = obs_windows, aes(y = confirm), size = 0.5) +
  geom_vline(
    data = tibble(forecast_date = forecast_dates),
    aes(xintercept = forecast_date),
    linetype = "dashed"
  ) +
  facet_wrap(~forecast_date, scales = "free", ncol = 1) +
  labs(
    x = "Date", y = "Reported cases", colour = "Model", fill = "Model",
    title = "Backtested forecasts by forecast date"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

During growth the models agree and all track the data.
Around the peak the three EpiNow2 models bend over with the observations because they all include susceptible depletion, while the ARIMA projects recent growth forward.
In the decline the recent downturn is already visible in the data, so all models project a fall and the gap between them narrows again.

::: callout-tip
## Take 5 minutes

Before reading on, predict which model you expect to score best overall, and at which forecast dates you expect the models to differ most.
Does your prediction change if we score on the natural scale versus the log scale?
:::

::: {.callout-note collapse="true"}
## Solution

The constant-RT fixed-population model should score best overall, because it matches the data-generating process of constant $R_t$ plus susceptible depletion.
The models should differ most around the peak and decline, where the changing trend is decisive.
During early growth the data already contain the stable trend, so the models converge.
Watch the constant-RT model as the peak passes, where a single fixed $R_t$ can be too rigid to track the turning trend.
:::

# Scoring

We merge the forecasts with the observed values and build a single `forecast_sample` object, keeping only out-of-sample horizons (`horizon > 0`).

```{r build-forecast-sample}
obs <- reported |> transmute(date, observed = confirm)

score_data <- all_forecasts |>
  filter(horizon > 0) |>
  left_join(obs, by = "date")

fc <- as_forecast_sample(
  score_data,
  forecast_unit = c("model", "forecast_date", "date", "horizon"),
  observed = "observed",
  predicted = "predicted",
  sample_id = "sample"
)
```

Reported cases span several orders of magnitude over an epidemic, so a score on the natural scale is dominated by the largest counts near the peak.
As in the [forecast evaluation session](forecast-evaluation.qmd), we score on both the natural and the log scale.
We use `transform_forecasts()` with `log_shift` to add log-transformed forecasts alongside the originals, which `score()` then scores together, labelling each with a `scale` column.

```{r score}
fc <- transform_forecasts(fc, fun = log_shift, offset = 1)
scores <- score(fc)
```

## Overall scores by model

```{r scores-by-model}
overall <- summarise_scores(scores, by = c("model", "scale")) |>
  as_tibble() |>
  mutate(across(where(is.numeric), \(x) signif(x, 3))) |>
  arrange(scale, crps)

overall |>
  select(model, scale, crps, bias, dispersion, underprediction, overprediction)
```

The results match the parsimony story we expected.
On the natural scale `constant_rt_fixed`, the model that matches the data-generating process, wins on CRPS.
The two random-walk models `rw_fixed` and `rw_estimated` sit in the middle.
Both pay for flexibility the data-generating process does not require, and `rw_estimated` pays a little more because the population is weakly identified from a single outbreak.
Their exact order can swap between runs under variational Bayes, but both are clearly worse than the correctly specified model and clearly better than the ARIMA.
The ARIMA has no epidemiological mechanism at all and is worst, because it cannot see the susceptible depletion that turns the curve over.
The correctly specified model wins, unnecessary parameters hurt, and removing mechanism entirely hurts most.
On the log scale the ranking is the same but the gaps are smaller, because log scoring down-weights the large peak counts that dominate the natural scale.
As we discussed in the [session on evaluating forecasts](forecast-evaluation.qmd), matching the data-generating process matters most when the mechanism shapes the forecast, here the susceptible depletion that bends the curve at the peak.
But `constant_rt_fixed` and `rw_fixed` use the true population size, information we would not have in the real world.
In practice you never know the population at risk precisely, and you cannot rule out the reproduction number changing over time as interventions, behaviour, and seasonality all shift it.
The model that estimates the population and lets $R_t$ vary, `rw_estimated`, is the one you would use.
The backtest shows the cost of that choice.
`rw_estimated` is competitive except near the peak, where the weakly identified population struggles, and its overall CRPS is worse than the fixed-population models.
That is the price of not assuming you know things you do not.

::: callout-tip
## Take 5 minutes

`constant_rt_fixed` wins on both CRPS and bias, so the correctly specified model is also the best-calibrated.
The ARIMA has the worst CRPS but relatively small bias, because its errors are symmetric.
It misses the magnitude of the peak but does not systematically over- or under-predict.
Recall from the [forecast evaluation session](forecast-evaluation.qmd) what CRPS and bias each measure.
Which metric should you trust when they disagree, and why?
:::

## Scores by forecast date

Breaking the CRPS down by forecast date shows how the ranking shifts across the epidemic's phases, on both the natural and log scales.

```{r scores-by-date}
by_date <- summarise_scores(
  scores, by = c("model", "forecast_date", "scale")
) |>
  as_tibble()

by_date |>
  arrange(forecast_date, scale, crps) |>
  select(forecast_date, model, scale, crps) |>
  mutate(crps = round(crps, 1))
```

We can also visualise this as a plot, with the observed case curve below for context.

```{r plot-scores-by-date}
scores_plot <- by_date |>
  ggplot(aes(x = forecast_date, y = crps, colour = model)) +
  geom_point() +
  geom_line() +
  facet_wrap(~scale, scales = "free_y") +
  labs(x = "Forecast date", y = "CRPS", colour = "Model") +
  theme_bw() +
  theme(legend.position = "bottom")

obs_plot <- reported |>
  filter(date >= min(forecast_dates), date <= max(forecast_dates) + 14) |>
  ggplot(aes(x = date, y = confirm)) +
  geom_col(fill = "#4292C6") +
  labs(x = "Date", y = "Reported cases") +
  theme_bw()

gridExtra::grid.arrange(scores_plot, obs_plot)
```

On the natural scale the pattern follows the epidemic phases.
During early growth (mid-February) `constant_rt_fixed` does best, because the correctly specified model wins when the trend is stable.
Around the peak (March) `rw_fixed` pulls ahead, because the random walk lets $R_t$ adapt as the data show the turnover.
During the decline (April) `constant_rt_fixed` does poorly, because its single $R_t$ is stuck too high.
`rw_fixed` and the ARIMA can adapt to the downward trend and do much better.
On the log scale the same pattern appears but is less pronounced, because log scoring down-weights the large counts near the peak that dominate the natural scale.
The natural scale is dominated by the peak counts, where the cost of a wrong mechanism is largest.
The log scale spreads the error more evenly across growth, peak, and decline.

## Scores by horizon

```{r scores-by-horizon}
by_horizon <- summarise_scores(
  scores, by = c("model", "horizon", "scale")
) |>
  as_tibble()

ggplot(by_horizon, aes(x = horizon, y = crps, colour = model)) +
  geom_line() +
  geom_point(size = 1) +
  facet_wrap(~scale, scales = "free_y") +
  labs(
    x = "Forecast horizon (days)", y = "Mean CRPS", colour = "Model",
    title = "Forecast error by horizon and scale"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

On the natural scale the gap between models opens at longer horizons, where the forecasts run into the peak.
On the log scale the differences are smaller and steadier, because log scoring down-weights the large counts that dominate the natural scale.

## Coverage

Interval coverage is not returned by `score()` for sample forecasts, so we compute it manually.
For each forecast we check whether the observation falls within the 50% and 90% prediction intervals, then average across all forecasts for each model.

```{r coverage}
coverage_data <- score_data |>
  group_by(model, forecast_date, date, horizon) |>
  summarise(
    in_50 = observed[1] >= quantile(predicted, 0.25, na.rm = TRUE) &
            observed[1] <= quantile(predicted, 0.75, na.rm = TRUE),
    in_90 = observed[1] >= quantile(predicted, 0.05, na.rm = TRUE) &
            observed[1] <= quantile(predicted, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  group_by(model) |>
  summarise(
    interval_coverage_50 = mean(in_50),
    interval_coverage_90 = mean(in_90)
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  arrange(model)

coverage_data
```

Well-calibrated models should hit approximately 50% and 90% coverage.
Variational Bayes gives approximate posteriors, so coverage may deviate from nominal levels.
The relative ranking between models is more informative than the absolute values.

# When the mechanism is wrong: a demonstration

The backtest showed that knowing the population matters.
The models with the fixed, correct population forecast the turnover well, while the model that estimates the population pays a cost near the peak.
Here we go further and deliberately give the model a population 10× too small, to see the cost of a wrong mechanism up close.
We fit two models at the peak forecast date, both with constant $R_t$ (`rw = 0`).
One uses the correct population and one uses a population 10× too small.

```{r mechanism-demo, warning = FALSE, message = FALSE}
demo_date <- as.Date("2023-03-18")
demo_train <- reported |> filter(date <= demo_date)

demo_correct <- fit_epinow2(
  rt_opts(pop = Fixed(pop_size), future = "latest", pop_period = "all"),
  demo_train
)

demo_wrong <- fit_epinow2(
  rt_opts(pop = Fixed(pop_size / 10), future = "latest", pop_period = "all"),
  demo_train
)

mechanism_compare <- bind_rows(
  get_predictions(demo_correct, format = "sample") |>
    as_tibble() |>
    filter(horizon > 0) |>
    mutate(model = "correct_pop"),
  get_predictions(demo_wrong, format = "sample") |>
    as_tibble() |>
    filter(horizon > 0) |>
    mutate(model = "wrong_pop")
) |>
  group_by(model, date) |>
  summarise(
    median = median(predicted, na.rm = TRUE),
    lower_90 = quantile(predicted, 0.05, na.rm = TRUE),
    upper_90 = quantile(predicted, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(mechanism_compare, aes(x = date, fill = model, colour = model)) +
  geom_ribbon(aes(ymin = lower_90, ymax = upper_90), alpha = 0.2) +
  geom_line(aes(y = median)) +
  geom_point(
    data = reported |> filter(date > demo_date - 14, date <= demo_date + 14),
    aes(x = date, y = confirm),
    inherit.aes = FALSE
  ) +
  geom_vline(xintercept = demo_date, linetype = "dashed") +
  scale_fill_manual(values = c("correct_pop" = "#4292C6", "wrong_pop" = "#E6550D")) +
  scale_colour_manual(values = c("correct_pop" = "#4292C6", "wrong_pop" = "#E6550D")) +
  labs(
    x = "Date", y = "Reported cases",
    title = "Correct vs wrong population size, forecast from 2023-03-18"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

The correct-population model bends over with the observations.
The wrong-population model is told the population is 10× too small, so susceptible depletion bites at the wrong time.
EpiNow2 even warns that the population is smaller than cumulative cases.
The only difference is one number, and it produces a systematically wrong forecast.
This is the cost of a wrong mechanism.
It is not noise but a directional error.

# Going further

## Challenge

-   Add a Gaussian process (`gp = gp_opts()`, from EpiNow2) instead of constant $R_t$. Does smoothing $R_t$ change the peak forecast?
-   Re-run the backtest with full MCMC (`stan_config <- stan_opts(samples = 1000, chains = 2, warmup = 300, method = 'sampling', backend = 'rstan')`). This takes around 20 minutes to refit every model. Do the rankings change with better-calibrated uncertainty?
-   Extend the forecast horizon to 28 days. Which model degrades fastest?

## Methods in practice

-   @abbottEpiNow2EstimateRealtime2025 on EpiNow2.
-   @bhattSemiMechanisticBayesianModeling2020 on renewal models in practice.
-   @gosticPracticalConsiderationsMeasuring2020 on practical $R_t$ estimation.

# Wrap up

-   Review what you've learned in this session with the [learning objectives](../reference/learning_objectives.qmd)
-   Share your [questions and thoughts](../reference/help.qmd)

## References

::: {#refs}
:::
