---
title: "Epidemiologically-motivated forecasting"
order: 5
bibliography: ../nfidd.bib
---

```{r echo=FALSE}
options(width=100)
```

# Introduction

In earlier sessions we built forecasts from purely statistical models such as ARIMA.
These models find patterns in reported counts and project those patterns forward.
They do not know anything about *how* infections arise, how they are transmitted, or why an epidemic must eventually decline.

In this session we add epidemiological mechanism to the forecast.
The key idea is the **renewal equation**, a model of incidence that is mathematically an autoregression whose coefficients are the generation-time distribution and whose scale is the reproduction number $R_t$ [@bhattSemiMechanisticBayesianModeling2020; @coriNewFrameworkSoftware2013].
This gives the forecast access to information that a pure time-series model cannot use.

We will account for the delay between infection and case report, and for susceptible depletion that bends the epidemic curve over at the peak.
We then backtest a set of models with different amounts of epidemiological mechanism, making forecasts from a series of past dates and scoring them against what happened next.
This asks a question about parsimony and identification.
Does the model that matches the data-generating process win, and do unnecessary parameters hurt?
Finally we deliberately break the mechanism with a wrong population size to see the cost of a wrong model up close.

::: callout-note
## `EpiNow2` is the tool, not the point
We fit these models with the [`EpiNow2`](https://epiforecasts.io/EpiNow2/) R package [@abbottEpiNow2EstimateRealtime2025].
`EpiNow2` combines the renewal equation, the delays from infection to report, and optional susceptible depletion, and fits them with [Stan](https://mc-stan.org/).
The same model structure fit with other software would give the same forecasts, so the focus here is on the structure rather than the package.
:::

## Slides

-   [Epidemiologically-motivated forecasting](slides/epi-motivated-forecasting.qmd)

## Objectives

The aim of this session is to introduce epidemiologically-motivated forecasting using renewal equations, to connect the renewal model to the autoregressive models from earlier sessions, and to backtest models with different levels of epidemiological mechanism against each other.

::: {.callout-note collapse="true"}
# Setup

## Source file

The source file of this session is located at `sessions/epi-motivated-forecasting.qmd`.

## Libraries used

In this session we will use the `EpiNow2` package for renewal-equation forecasting, the `fable` package for the ARIMA baseline, the `dplyr`, `tidyr`, `tibble`, and `purrr` packages for data wrangling, the `ggplot2` and `gridExtra` packages for plotting, and the `scoringutils` package for forecast evaluation.

```{r libraries, message = FALSE}
library("nfidd.forecasting")
library("EpiNow2")
library("fable")
library("dplyr")
library("tidyr")
library("tibble")
library("purrr")
library("ggplot2")
library("gridExtra")
library("scoringutils")
theme_set(theme_bw())
```

::: callout-tip
The best way to interact with the material is via the [Visual Editor](https://docs.posit.co/ide/user/ide/guide/documents/visual-editor.html) of RStudio.
:::

## Initialisation

We set a random seed for reproducibility.
Setting this ensures that you should get exactly the same results on your computer as we do.
This is not strictly necessary but will help us talk about the models.

```{r}
set.seed(7321)
```

We also define the fourth-root transformation used in earlier sessions, so that our ARIMA baseline is the same model that you are already familiar with.

```{r}
fourth_root <- function(x) x^0.25
inv_fourth_root <- function(x) x^4
my_fourth_root <- new_transformation(fourth_root, inv_fourth_root)
```

We set the inference configuration used for the `EpiNow2` fits in this session.
We use Markov-chain Monte Carlo (MCMC) sampling, which gives the most honest posterior; you could increase the warmup and sampling iterations for your own runs.

```{r}
stan_config <- stan_opts(
  method = "sampling", backend = "rstan",
  chains = 2, warmup = 300, samples = 1000
)
```
:::

# From autoregression to the renewal equation

An ARIMA model predicts the next value of a time series as a weighted sum of its past values.
A renewal equation has the same autoregressive shape, but every piece is given epidemiological meaning.

:::: {.columns}
::: {.column width="48%"}
#### ARIMA

$$ y_t = \sum_s \phi_s\, y_{t-s} + \varepsilon_t $$

- Coefficients $\phi_s$ are free, estimated from the data
- $\varepsilon_t$ is an additive noise term
- No epidemiological meaning
:::

::: {.column width="48%"}
#### Renewal

$$ I_t = \sum_s R_t\, w_s\, I_{t-s} + \iota_t $$

- Coefficient on each past value is $R_t\, w_s$, the reproduction number times the generation-time weight
- $\iota_t$ sits where the ARIMA has $\varepsilon_t$, and is imported infections, set to zero here
- Every term is an epidemiological quantity
:::
::::

Here $I_t$ is the number of new infections, $R_t$ is the reproduction number, and $w_s$ is the **generation-time distribution**, the probability that an infection caused $s$ days ago transmits now [@wallingaDifferentEpidemicCurves2004].
Written this way the two models have the same structure: the current value is a weighted sum of past values plus an additive term.
The ARIMA weights are free, while the renewal weights are the reproduction number times the generation-time distribution, so the epidemiology sets the coefficients.
Where the ARIMA has a noise term $\varepsilon_t$, the renewal has an imports term $\iota_t$ for infections arriving from outside the population, which we set to zero here.
This is the deterministic renewal equation; when we fit it we add an observation model (negative binomial) on top to account for noise in the reported counts.
The generation-time weights $w_s$ are not free but must form a valid probability distribution, and are typically estimated from data on serial intervals or generation times, not from the incidence curve itself.

# Simulating an outbreak

We simulate a single outbreak to forecast, using `simulate_infections()` from `EpiNow2`.
We set a constant reproduction number above 1 and a fixed population size, which produces an epidemic that grows, peaks, and then declines as susceptibles are depleted.
Setting `pop_period = "all"` applies the depletion across the whole simulated period, so the turnover is built into the data-generating process.
The generation time and delays are the example distributions supplied with `EpiNow2`, fixed to point values with `fix_parameters()`.
The observations are negative binomial with a fixed, moderate overdispersion, which is the observation model our forecasts will also use.

The simulation and the backtest that follows are run ahead of time in `data-raw/epi_forecasts.R` and saved with the `nfidd.forecasting` package, so the session renders quickly.
We define the fixed distributions we need for the live fits, show the simulation code for reference, and load the saved series.

```{r sim-params}
gen_time <- fix_parameters(example_generation_time)
delays <- fix_parameters(example_incubation_period + example_reporting_delay)
pop_size <- 100000
```

```{r simulate-code, eval = FALSE}
dates <- seq.Date(as.Date("2023-01-01"), by = "day", length.out = 170)
R <- data.frame(date = dates, R = 1.25)

sim <- simulate_infections(
  R = R,
  initial_infections = 20,
  generation_time = gt_opts(gen_time),
  delays = delay_opts(delays),
  obs = obs_opts(family = "negbin", dispersion = Fixed(0.1)),
  pop = Fixed(pop_size),
  pop_period = "all"
)

reported <- sim |>
  as_tibble() |>
  filter(variable == "reported_cases") |>
  transmute(date, confirm = value)
```

```{r load-reported}
data(epi_reported)
reported <- epi_reported
```

The simulated series of reported cases rises to a peak in late March before declining.

```{r plot-outbreak}
ggplot(reported, aes(x = date, y = confirm)) +
  geom_col(fill = "#4292C6") +
  labs(x = "Date", y = "Reported cases", title = "Simulated outbreak") +
  theme_bw()
```

The decline happens because susceptibles are running out.
The population at risk is finite, so once enough people have been infected the effective reproduction number drops below 1 even though the basic reproduction number $R_0$ we set is still 1.25.

# A single forecast

To forecast we truncate the series at a forecast date and fit only to the data available up to that point.
We choose a date shortly before the peak, where the epidemic is still growing and the turnover has not yet been observed.

```{r truncate}
forecast_date <- as.Date("2023-03-18")
train <- reported |> filter(date <= forecast_date)
```

Every renewal fit in this session uses `fit_epinow2()` from the `nfidd.forecasting` package.
It holds the generation time and delays fixed at the values used to simulate the data, turns off the day-of-week effect because the simulation has none, and uses a negative binomial observation model to match the simulation.
The same helper runs the backtest in `data-raw/epi_forecasts.R`, and you can reuse it for your own real-time estimates.

Under the hood `fit_epinow2()` calls `estimate_infections()`, which combines the renewal infection model, the delay convolution, and the observation model in a single Bayesian fit.

![The `EpiNow2` model architecture. Diagram from the [EpiNow2 model overview](https://epiforecasts.io/EpiNow2/articles/estimate_infections.html).](slides/figures/epinow2-architecture.png)

::: callout-tip
## Prototyping with faster inference

The fits here use MCMC, defined in `stan_config` above and passed to `fit_epinow2()`.
While developing your own analysis you can swap in a faster approximate method by changing `stan_config`, for example variational Bayes:

```r
stan_config <- stan_opts(method = "vb", backend = "rstan", samples = 1000)
```

Variational Bayes returns in seconds but gives an approximate posterior, so use MCMC for final results.
:::

We also define a helper that fits the ARIMA baseline with `fable` and returns sample-level predictions.

```{r arima-helper}
forecast_arima_samples <- function(train, horizon = 14, times = 1000) {
  fit <- train |>
    as_tsibble(index = date) |>
    model(ar2 = ARIMA(my_fourth_root(confirm)))
  generate(fit, h = horizon, times = times) |>
    as_tibble() |>
    transmute(
      forecast_date = max(train$date),
      date,
      horizon = as.integer(date - max(train$date)),
      sample = as.integer(as.factor(.rep)),
      predicted = pmax(.sim, 0),
      model = "arima"
    )
}
```

We fit the EpiNow2 default renewal model, a weekly random walk on $R_t$ with no susceptible depletion, and the ARIMA baseline.
Both use a 14-day forecast horizon.

```{r single-forecast, warning = FALSE, message = FALSE}
en2_default <- fit_epinow2(rt_opts(rw = 7), train, stan = stan_config)
arima_default <- forecast_arima_samples(train)
```

We plot both forecasts against the data that were later observed, on the natural and the log scale.
The shaded ribbons are the 50% and 90% credible intervals, the lines are the posterior medians, and the points are the observations.
A renewal model with no susceptible depletion can grow explosively past the peak, so on the natural scale we cap the axis to keep the plot readable; the log scale shows the full range.

```{r plot-single-forecast, fig.width = 10, fig.height = 5, warning = FALSE, message = FALSE}
single_fc <- bind_rows(
  get_predictions(en2_default, format = "sample") |>
    as_tibble() |>
    filter(horizon > 0) |>
    transmute(date, sample = as.integer(sample), predicted,
              model = "epi_default"),
  arima_default |> transmute(date, sample, predicted, model)
) |>
  filter(!is.na(predicted))

# 30%, 60% and 90% prediction intervals
band_summary <- function(df, group_vars) {
  df |>
    group_by(across(all_of(group_vars))) |>
    summarise(
      lower_30 = quantile(predicted, 0.35, na.rm = TRUE),
      upper_30 = quantile(predicted, 0.65, na.rm = TRUE),
      lower_60 = quantile(predicted, 0.20, na.rm = TRUE),
      upper_60 = quantile(predicted, 0.80, na.rm = TRUE),
      lower_90 = quantile(predicted, 0.05, na.rm = TRUE),
      upper_90 = quantile(predicted, 0.95, na.rm = TRUE),
      .groups = "drop"
    )
}

# draw the three nested intervals as ribbons, no median line
add_bands <- function(g) {
  g +
    geom_ribbon(aes(ymin = lower_90, ymax = upper_90, fill = model),
                alpha = 0.15) +
    geom_ribbon(aes(ymin = lower_60, ymax = upper_60, fill = model),
                alpha = 0.25) +
    geom_ribbon(aes(ymin = lower_30, ymax = upper_30, fill = model),
                alpha = 0.40)
}

single_sum <- band_summary(single_fc, c("model", "date"))

obs_window <- reported |>
  filter(date > forecast_date - 14, date <= forecast_date + 14)

plot_single <- function(log_scale = FALSE) {
  g <- add_bands(ggplot(single_sum, aes(x = date))) +
    geom_point(data = obs_window, aes(x = date, y = confirm),
               inherit.aes = FALSE, size = 1.2) +
    geom_vline(xintercept = forecast_date, linetype = "dashed") +
    scale_fill_manual(
      values = c("epi_default" = "#4292C6", "arima" = "#E6550D")) +
    labs(x = "Date", y = "Reported cases", fill = "Model") +
    theme_bw() +
    theme(legend.position = "bottom")
  if (log_scale) {
    g + scale_y_log10(limits = c(1, NA)) + labs(title = "Log scale")
  } else {
    g + coord_cartesian(ylim = c(0, max(obs_window$confirm) * 1.8)) +
      labs(title = "Natural scale")
  }
}

gridExtra::grid.arrange(plot_single(FALSE), plot_single(TRUE), ncol = 2)
```

Neither model knows that the epidemic is about to turn over.
The ARIMA projects the recent trend forward, and the renewal default does the same because it has no mechanism for susceptible depletion.
Both over-shoot the peak.
This foreshadows the backtest, where we give the other EpiNow2 models the ability to see the turnover coming.

::: callout-tip
## Take 5 minutes

Look at the two forecasts against the held-out observations, on both scales.
Which model looks closer over the first few days, and which drifts further off as the horizon grows?
What single piece of information would either model need to predict the turning point?
:::

::: {.callout-note collapse="true"}
## Solution

Over the first few days both forecasts sit close to the observations, and there is little to choose between them.
As the horizon grows both drift above the truth and over-shoot the peak, because neither has any mechanism to stop growth.
The renewal default grows fastest, since it projects the recent reproduction number forward with nothing to bend it down.
To predict the turning point either model would need to know that the susceptible pool is running out, which means knowing the population at risk and how much of it has already been infected.
:::

# The models we will compare

We compare five models that differ in how much epidemiological mechanism and flexibility they include.
We have already met two of them, the ARIMA baseline and the EpiNow2 default renewal model.
This is a story about parsimony and identification.
The data-generating process has constant $R_t$ and a known population, so the model that matches it should do well.
Unnecessary flexibility (a random walk on $R_t$) and an unidentified parameter (estimating the population) should carry a cost.

- **arima**. `ARIMA(my_fourth_root(confirm))` with auto-ARIMA, so fable selects the order automatically from the data. It may select different orders at different forecast dates, which is a feature of a fair statistical baseline rather than a bug.
- **default**. A renewal model with a weekly random walk on $R_t$ and no susceptible depletion, the EpiNow2 default we used for the single forecast. It knows about transmission and delays but has nothing to bend the curve at the peak.
- **constant_rt_fixed**. Constant reproduction number (the default `rw = 0`) with the *true* population size. This matches the data-generating process of constant $R_t$ plus susceptible depletion.
- **rw_fixed**. Adds a weekly random walk on $R_t$ (`rw = 7`) with the true population. The random walk is extra flexibility the data-generating process does not require.
- **rw_estimated**. Weekly random walk on $R_t$ with an estimated population (`Normal(mean = pop_size, sd = pop_size / 2)`). The population is weakly identified from a single outbreak, so estimating it adds uncertainty.

The four EpiNow2 models share the same renewal structure and differ only in the `rt_opts()` arguments.
The ARIMA baseline is fit separately with `fable`, using the same fourth-root transformation as in earlier sessions.

::: callout-note
## What is a random walk on $R_t$?
We met the random walk in earlier sessions as a baseline that carries the last value forward with added noise.
Here we place the random walk on $R_t$ rather than on the cases, so $R_t$ drifts from week to week instead of being held constant.
This is a common default for $R_t$ estimation, because it lets the reproduction number change over time without assuming a particular shape.
:::

The wrong-population model is deliberately left out of the backtest.
It appears later as a standalone demonstration.

```{r model-specs}
epi_models <- list(
  default           = rt_opts(rw = 7),
  constant_rt_fixed = rt_opts(pop = Fixed(pop_size), future = "latest", pop_period = "all"),
  rw_fixed          = rt_opts(rw = 7, pop = Fixed(pop_size), future = "latest", pop_period = "all"),
  rw_estimated      = rt_opts(rw = 7, pop = Normal(mean = pop_size, sd = pop_size / 2), future = "latest", pop_period = "all")
)
```

::: callout-note
## What the `rt_opts()` settings mean

- `rw = 7` puts a weekly random walk on $R_t$; leave it out for a constant $R_t$ (`rw = 0`).
- `pop = Fixed(...)` or `Normal(...)` sets the population at risk, which turns on susceptible depletion; without it there is no depletion.
- `future = "latest"` carries the last estimated $R_t$ forward across the forecast horizon.
- `pop_period = "all"` applies the depletion across the whole period.
:::

# Backtesting across forecast dates

We backtest each model from 12 forecast dates, one week apart, spanning the growth phase, the peak, and the decline.
Contrasting phases matters because susceptible depletion only shapes the forecast near and after the peak.
Early in the growth phase it is almost irrelevant.

```{r forecast-dates}
forecast_dates <- seq.Date(
  from = as.Date("2023-02-01"), to = as.Date("2023-04-19"), by = "week"
)
length(forecast_dates)
```

At each forecast date we fit all four EpiNow2 models and the ARIMA baseline with the helpers defined above, and combine the sample-level predictions into a single tibble.
This backtest fits 60 models, so we run it ahead of time in `data-raw/epi_forecasts.R` and load the saved result; the code is shown for reference.

```{r backtest-code, eval = FALSE}
backtest_one <- function(fdate) {
  train <- reported |> filter(date <= fdate)
  en2 <- map_dfr(names(epi_models), function(m) {
    fit <- fit_epinow2(epi_models[[m]], train, stan = stan_config)
    get_predictions(fit, format = "sample") |>
      as_tibble() |>
      filter(horizon > 0) |>
      transmute(
        forecast_date = fdate, date, horizon,
        sample = as.integer(sample), predicted, model = m
      )
  })
  arm <- forecast_arima_samples(train)
  bind_rows(en2, arm)
}

all_forecasts <- map_dfr(forecast_dates, backtest_one) |>
  filter(!is.na(predicted))
```

```{r load-backtest}
data(epi_forecasts)
all_forecasts <- epi_forecasts
```

## Visualising the backtested forecasts

Before scoring, we look at the forecast trajectories themselves.
Each panel is a forecast date, with the 50% and 90% credible intervals as ribbons, the posterior medians as lines, and the observed data as points.
The no-depletion models can grow explosively, so we give each panel its own y axis and show the trajectories on both the natural and the log scale.

```{r plot-backtests}
trajectory_summary <- band_summary(
  all_forecasts |> filter(horizon > 0, !is.na(predicted)),
  c("model", "forecast_date", "date")
)

obs_windows <- map_dfr(forecast_dates, function(fd) {
  reported |>
    filter(date >= fd - 14, date <= fd + 14) |>
    mutate(forecast_date = fd)
})

# each panel gets its own y axis; on the natural scale we clip the ribbons
# to twice that panel's observed peak so the exploding no-depletion models
# do not dominate the axis
panel_caps <- obs_windows |>
  group_by(forecast_date) |>
  summarise(cap = max(confirm) * 2, .groups = "drop")

plot_trajectories <- function(log_scale = FALSE) {
  dat <- trajectory_summary
  if (!log_scale) {
    dat <- dat |>
      left_join(panel_caps, by = "forecast_date") |>
      mutate(across(
        c(lower_30, upper_30, lower_60, upper_60, lower_90, upper_90),
        \(x) pmin(x, cap)
      ))
  }
  add_bands(ggplot(dat, aes(x = date))) +
    geom_point(data = obs_windows, aes(y = confirm), size = 0.5) +
    geom_vline(
      data = tibble(forecast_date = forecast_dates),
      aes(xintercept = forecast_date), linetype = "dashed", linewidth = 0.3
    ) +
    facet_wrap(~forecast_date, scales = "free", ncol = 3) +
    labs(x = "Date", y = "Reported cases", fill = "Model") +
    theme_bw() +
    theme(legend.position = "bottom") +
    (if (log_scale) {
      list(scale_y_log10(limits = c(1, NA)), labs(title = "Log scale"))
    } else {
      labs(title = "Natural scale")
    })
}
```

```{r plot-backtests-natural, fig.width = 11, fig.height = 11}
plot_trajectories(FALSE)
```

```{r plot-backtests-log, fig.width = 11, fig.height = 11, warning = FALSE, message = FALSE}
plot_trajectories(TRUE)
```

During growth the models agree and track the data.
Around the peak the depletion models (`constant_rt_fixed`, `rw_fixed`, `rw_estimated`) bend over with the observations, while the `default` renewal and the ARIMA project growth forward and over-shoot.
In the decline the downturn is already visible in the data, so all models project a fall and the gap between them narrows again.

::: callout-tip
## Take 5 minutes

Before reading on, predict which model you expect to score best overall, and at which forecast dates you expect the models to differ most.
Does your prediction change if we score on the natural scale versus the log scale?
:::

::: {.callout-note collapse="true"}
## Solution

The depletion models should score best, because they can bend the curve at the peak, and the model that matches the data-generating process, `constant_rt_fixed`, should be the best of them.
The models should differ most around the peak, where a mechanism for the turnover matters most.
During early growth the data already contain the stable trend, so the models converge.
The `default` renewal and the ARIMA should do worst on the natural scale, where over-shooting the large peak counts is heavily penalised.
:::

# Scoring

We merge the forecasts with the observed values, keeping only out-of-sample horizons (`horizon > 0`).
The ARIMA returns 1000 samples while the EpiNow2 fits occasionally drop one, so we take the same number of samples from every forecast before scoring, which keeps `score()` from warning about uneven sample counts.

```{r build-forecast-sample}
obs <- reported |> transmute(date, observed = confirm)

score_data <- all_forecasts |>
  filter(horizon > 0) |>
  left_join(obs, by = "date")

common_n <- score_data |>
  count(model, forecast_date, date, horizon) |>
  pull(n) |>
  min()

score_data <- score_data |>
  group_by(model, forecast_date, date, horizon) |>
  slice_sample(n = common_n) |>
  mutate(sample = row_number()) |>
  ungroup()

fc <- as_forecast_sample(
  score_data,
  forecast_unit = c("model", "forecast_date", "date", "horizon"),
  observed = "observed",
  predicted = "predicted",
  sample_id = "sample"
)
```

Reported cases span several orders of magnitude over an epidemic, so a score on the natural scale is dominated by the largest counts near the peak.
As in the [forecast evaluation session](forecast-evaluation.qmd), we score on both the natural and the log scale.
We use `transform_forecasts()` with `log_shift` to add log-transformed forecasts alongside the originals, which `score()` then scores together, labelling each with a `scale` column.

```{r score}
fc <- transform_forecasts(fc, fun = log_shift, offset = 1)
scores <- score(fc)
```

## Overall scores by model

```{r scores-by-model}
overall <- summarise_scores(scores, by = c("model", "scale")) |>
  as_tibble() |>
  mutate(across(where(is.numeric), \(x) signif(x, 3))) |>
  arrange(scale, crps)

overall |>
  select(model, scale, crps, bias, dispersion, underprediction, overprediction) |>
  knitr::kable()
```

::: callout-tip
## Take 5 minutes

Read the score table before reading on.
On each scale, which model has the best (lowest) CRPS, and which the worst?
What do the `bias`, `overprediction`, and `underprediction` columns say about *how* each model fails?
How does the ranking change between the natural and the log scale, and why?
:::

::: {.callout-note collapse="true"}
## Solution

- `constant_rt_fixed`, the model that matches the data-generating process, has the best CRPS on both scales, with `rw_fixed` just behind.
- The `default` renewal and the ARIMA are far worse, and their error is dominated by overprediction because they sail past the peak. On the natural scale the `default` is worst; on the log scale the ARIMA is worst, because log scoring reweights where the errors fall.
- `rw_estimated` sits in between, because estimating the population from a single outbreak adds uncertainty, so it pays a little over the fixed-population models.
- On the log scale the gaps shrink, because log scoring down-weights the large peak counts that dominate the natural scale, but the depletion models still lead.
- The fixed-population models use the true population, which you would not know in the real world, so `rw_estimated` is closer to what you would actually run.
:::

## Scores by forecast date

Breaking the CRPS down by forecast date shows how the ranking shifts across the epidemic's phases.
The table is hard to read across models and dates, so we go straight to the plot, with the observed case curve below for context.

```{r scores-by-date}
by_date <- summarise_scores(
  scores, by = c("model", "forecast_date", "scale")
) |>
  as_tibble()

scores_plot <- by_date |>
  ggplot(aes(x = forecast_date, y = crps, colour = model)) +
  geom_point() +
  geom_line() +
  facet_wrap(~scale, scales = "free_y") +
  labs(x = "Forecast date", y = "CRPS", colour = "Model") +
  theme_bw() +
  theme(legend.position = "bottom")

obs_plot <- reported |>
  filter(date >= min(forecast_dates), date <= max(forecast_dates) + 14) |>
  ggplot(aes(x = date, y = confirm)) +
  geom_col(fill = "#4292C6") +
  labs(x = "Date", y = "Reported cases") +
  theme_bw()

gridExtra::grid.arrange(scores_plot, obs_plot, heights = c(2, 1))
```

The models are close early in growth, because the data already contain the stable trend.
Around the peak the gap is largest: `constant_rt_fixed` and `rw_fixed` stay accurate while the `default` renewal and the ARIMA blow up by over-shooting, and `rw_estimated` struggles a little because its population is weakly identified.
Into the decline the downturn is visible in the data and the models come back together.
On the log scale the same pattern appears but is less pronounced, because log scoring down-weights the large peak counts that dominate the natural scale.

## Scores by horizon

```{r scores-by-horizon}
by_horizon <- summarise_scores(
  scores, by = c("model", "horizon", "scale")
) |>
  as_tibble()

ggplot(by_horizon, aes(x = horizon, y = crps, colour = model)) +
  geom_line() +
  geom_point(size = 1) +
  facet_wrap(~scale, scales = "free_y") +
  labs(
    x = "Forecast horizon (days)", y = "Mean CRPS", colour = "Model",
    title = "Forecast error by horizon and scale"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

On the natural scale the gap between models opens at longer horizons, where the forecasts run into the peak.
On the log scale the differences are smaller and steadier, because log scoring down-weights the large counts that dominate the natural scale.

## Coverage

Coverage asks how often the observations fall inside the forecast intervals.
A well-calibrated 50% interval should contain the observation about half the time, and a 90% interval about nine times in ten.
We convert the sample forecasts to quantiles with `as_forecast_quantile()`, and then `score()` and `summarise_scores()` return interval coverage directly.

```{r coverage}
coverage <- score_data |>
  as_forecast_sample(
    forecast_unit = c("model", "forecast_date", "date", "horizon"),
    observed = "observed", predicted = "predicted", sample_id = "sample"
  ) |>
  as_forecast_quantile(probs = c(0.05, 0.25, 0.5, 0.75, 0.95)) |>
  score() |>
  summarise_scores(by = "model") |>
  as_tibble() |>
  transmute(
    model,
    interval_coverage_50 = round(interval_coverage_50, 2),
    interval_coverage_90 = round(interval_coverage_90, 2)
  ) |>
  arrange(model)

coverage |> knitr::kable()
```

::: callout-tip
## Take 5 minutes

Compare each model's 50% and 90% coverage against the nominal 0.5 and 0.9.
Which models are over-confident, with intervals too narrow to reach nominal coverage, and which are about right?
:::

::: {.callout-note collapse="true"}
## Solution

- The depletion models are close to nominal: `constant_rt_fixed` covers about 0.58 and 0.91, `rw_fixed` 0.61 and 0.90, and `rw_estimated` 0.61 and 0.98.
- The ARIMA (about 0.18 and 0.39) and the `default` renewal (about 0.38 and 0.62) are badly over-confident: their intervals are far too narrow to contain the observations around the peak they miss.
- Coverage is a joint check on the median and the spread: a model can have reasonable coverage by being wide as well as by being accurate, so read it alongside the CRPS rather than on its own.
:::

# When the modeller is misinformed: a demonstration

The backtest showed that knowing the population matters.
The model is only ever as good as the information we give it, so here we misinform it, the kind of mistake a modeller could make in practice.
We fit two models at the peak forecast date, both with constant $R_t$ and susceptible depletion.
One is given the true population.
The other is given a prior centred on a population ten times too small, `Normal(mean = pop_size / 10, sd = pop_size / 20)`, an honest but mistaken guess by whoever set the model up.

```{r mechanism-demo, results = "hide", warning = FALSE, message = FALSE}
demo_date <- as.Date("2023-03-18")
demo_train <- reported |> filter(date <= demo_date)

demo_correct <- fit_epinow2(
  rt_opts(pop = Fixed(pop_size), future = "latest", pop_period = "all"),
  demo_train, stan = stan_config
)

demo_wrong <- fit_epinow2(
  rt_opts(pop = Normal(mean = pop_size / 10, sd = pop_size / 20),
          future = "latest", pop_period = "all"),
  demo_train, stan = stan_config
)

mechanism_compare <- bind_rows(
  get_predictions(demo_correct, format = "sample") |>
    as_tibble() |> filter(horizon > 0) |> mutate(model = "correct_pop"),
  get_predictions(demo_wrong, format = "sample") |>
    as_tibble() |> filter(horizon > 0) |> mutate(model = "wrong_pop_prior")
) |>
  filter(!is.na(predicted)) |>
  group_by(model, date) |>
  summarise(
    median = median(predicted, na.rm = TRUE),
    lower_90 = quantile(predicted, 0.05, na.rm = TRUE),
    upper_90 = quantile(predicted, 0.95, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r plot-mechanism-demo}
ggplot(mechanism_compare, aes(x = date, fill = model, colour = model)) +
  geom_ribbon(aes(ymin = lower_90, ymax = upper_90), alpha = 0.2,
              colour = NA) +
  geom_line(aes(y = median)) +
  geom_point(
    data = reported |> filter(date > demo_date - 14, date <= demo_date + 14),
    aes(x = date, y = confirm), inherit.aes = FALSE
  ) +
  geom_vline(xintercept = demo_date, linetype = "dashed") +
  scale_fill_manual(
    values = c("correct_pop" = "#4292C6", "wrong_pop_prior" = "#E6550D")) +
  scale_colour_manual(
    values = c("correct_pop" = "#4292C6", "wrong_pop_prior" = "#E6550D")) +
  labs(x = "Date", y = "Reported cases", fill = "Model", colour = "Model",
       title = "Correct population vs a wrong population prior") +
  theme_bw() +
  theme(legend.position = "bottom")
```

::: callout-tip
## Take 5 minutes

Compare the two forecasts against the observations after the forecast date.
What does the wrong population prior do to the timing of the turnover, and in which direction is the error?
Is the wrong forecast simply noisier, or is it wrong in a systematic way?
:::

::: {.callout-note collapse="true"}
## Solution

- The correct-population model bends over with the observations.
- The wrong prior tells the model the susceptible pool is far smaller than it is, so depletion takes hold too early and the forecast turns over before the real peak, under-shooting the truth.
- The error is systematic and directional, not noise: a wrong mechanism sends the forecast the wrong way rather than just widening the intervals.
:::

# Going further

## Challenge

-   Change the simulation and see how the rankings move. Try a different reproduction number, a different population size, or a reproduction number that varies over time (`R` as a time series in `simulate_infections()`). Note that a time-varying `R` in the simulation is the effective reproduction number for a wholly susceptible population, before susceptible depletion is applied on top.
-   Give the ARIMA baseline some mechanism. Build a susceptible-fraction covariate, for example `1 - cumsum(confirm) / pop_size`, and add it as a regressor with `ARIMA(my_fourth_root(confirm) ~ susceptible_fraction)`. Because the covariate depends on the cases, you would forecast one step ahead at a time, updating the cumulative cases with each prediction. Does a statistical model with a depletion covariate close the gap to the renewal models?
-   Add a Gaussian process (`gp = gp_opts()`, from EpiNow2) instead of a random walk on $R_t$. Does smoothing $R_t$ change the peak forecast?
-   Swap the inference to variational Bayes or pathfinder by changing the single `stan_config` line. How much faster is the backtest, and does the coverage change?
-   Extend the forecast horizon to 28 days. Which model degrades fastest?

## Methods in practice

-   @abbottEpiNow2EstimateRealtime2025 on EpiNow2.
-   @bhattSemiMechanisticBayesianModeling2020 on renewal models in practice.
-   @gosticPracticalConsiderationsMeasuring2020 on practical $R_t$ estimation.

# Wrap up

-   Review what you've learned in this session with the [learning objectives](../reference/learning_objectives.qmd)
-   Share your [questions and thoughts](../reference/help.qmd)

## References

::: {#refs}
:::
