---
title: "Epi-motivated forecasting"
order: 5
bibliography: ../nfidd.bib
---

```{r echo=FALSE}
options(width=100)
```

# Introduction

In earlier sessions we built forecasts from purely statistical models — ARIMA, baselines, and ensembles.
These models find patterns in reported counts and project those patterns forward.
They do not know anything about *how* infections arise, how they are transmitted, or why an epidemic must eventually decline.

In this session we add epidemiological mechanism to the forecast.
The key idea is the **renewal equation**: a model of incidence that is, mathematically, an autoregression whose coefficients are the generation-time distribution and whose scale is the reproduction number $R_t$ [@bhattSemiMechanisticBayesianModeling2020; @coriNewFrameworkSoftware2013].
This gives the forecast access to information that a pure time-series model cannot use — for example, that exponential growth must end when susceptibles run out.

We will also account for the delay between infection and case report, and for susceptible depletion that bends the epidemic curve over at the peak.
To see why mechanism matters, we will deliberately give one model the *wrong* population size and watch its forecast fail in a predictable direction.
Throughout, we use the EpiNow2 R package as our tool [@abbottEpiNow2EstimateRealtime2025]; the focus is on the model structure, not the software.

## Slides

-   [Epi-motivated forecasting](slides/epi-motivated-forecasting.qmd)

## Objectives

The aim of this session is to introduce epidemiologically-motivated forecasting using renewal equations, to connect the renewal model to the autoregressive models from earlier sessions, and to backtest models with different levels of epidemiological mechanism against each other.

::: {.callout-note collapse="true"}
# Setup

## Source file

The source file of this session is located at `sessions/epi-motivated-forecasting.qmd`.

## Libraries used

In this session we will use the `EpiNow2` package for renewal-equation forecasting, the `fable` package for the ARIMA baseline, the `dplyr`, `tidyr`, `tibble`, and `purrr` packages for data wrangling, the `ggplot2` and `gridExtra` packages for plotting, and the `scoringutils` package for forecast evaluation.

```{r libraries, message = FALSE}
library("nfidd.forecasting")
library("EpiNow2")
library("fable")
library("dplyr")
library("tidyr")
library("tibble")
library("purrr")
library("ggplot2")
library("gridExtra")
library("scoringutils")
theme_set(theme_bw())
```

::: callout-tip
The best way to interact with the material is via the [Visual Editor](https://docs.posit.co/ide/user/ide/guide/documents/visual-editor.html) of RStudio.
:::

## Initialisation

We set a random seed for reproducibility.
Setting this ensures that you should get exactly the same results on your computer as we do.
This is not strictly necessary but will help us talk about the models.

```{r}
set.seed(7321)
```

We also define the fourth-root transformation used in earlier sessions, so that our ARIMA baseline is the same model family students already know.

```{r}
fourth_root <- function(x) x^0.25
inv_fourth_root <- function(x) x^4
my_fourth_root <- new_transformation(fourth_root, inv_fourth_root)
```

We also set the inference configuration that every EpiNow2 fit in this session will use, so we can change it in one place if we want different inference later.

```{r}
stan_config <- stan_opts(method = "vb", backend = "rstan", samples = 1000)
```
:::

::: {.callout-note collapse="true"}
## Inference in this session: variational Bayes, and how to switch to pathfinder or MCMC

EpiNow2 fits a Bayesian model in Stan, and we refit it dozens of times across forecast dates and models.
Full Markov-chain Monte Carlo (MCMC) sampling gives the most honest posterior but takes ~1–2 minutes per fit, which would make the backtest take over an hour.
Variational Bayes (`vb`) gives an approximate posterior in seconds via the `rstan` backend that ships with EpiNow2, so you do not need to install anything extra — the right trade-off for a teaching backtest where we care about relative model behaviour, not calibrated uncertainty.

If you have the `cmdstanr` package and a CmdStan install, pathfinder usually gives a better approximation than `vb`.
Switch by changing the single `stan_config` line above:

```r
stan_config <- stan_opts(method = "pathfinder", backend = "cmdstanr")
```

For full MCMC (best-calibrated, slowest):

```r
stan_config <- stan_opts(samples = 1000, chains = 2, method = "sampling", backend = "rstan")
```

Every fit in the session picks up the change automatically.
:::

# From autoregression to the renewal equation

An ARIMA model, as we saw in earlier sessions, predicts the next value of a time series as a weighted sum of past values:
$$ y_t = \sum_s \phi_s\, y_{t-s} + \varepsilon_t $$
The weights $\phi_s$ are free parameters, estimated from the data with no constraints beyond stationarity.

A renewal equation has the same autoregressive shape, but every piece is given epidemiological meaning:
$$ I_t \sim \mathrm{Poisson}\!\left(R_t \sum_s w_s\, I_{t-s}\right) $$
Here $I_t$ is the number of new infections, $R_t$ is the reproduction number, and $w_s$ is the **generation-time distribution** — the probability that an infection caused $s$ days ago transmits now [@wallingaDifferentEpidemicCurves2004].
The weights are not free: they must form a valid probability distribution, and they are typically estimated from data on serial intervals or generation times, not from the incidence curve itself.

:::: {.columns}
::: {.column width="48%"}
#### ARIMA

$$ y_t = \sum_s \phi_s\, y_{t-s} + \varepsilon_t $$

- Weights $\phi_s$: free, estimated from the data
- Scale: implicit in the weights
- No epidemiological meaning
:::

::: {.column width="48%"}
#### Renewal

$$ I_t \sim \mathrm{Poisson}\!\left(R_t \sum_s w_s\, I_{t-s}\right) $$

- Weights $w_s$: generation-time PMF
- Scale: $R_t$ (reproduction number)
- Every term is an epidemiological quantity
:::
::::

This is the conceptual bridge: **a renewal equation is an epidemiologically-informed autoregression**.
The structure is familiar, but the coefficients now carry mechanistic meaning that constrains what the model can say about the future.

# Simulating an outbreak

We simulate a single outbreak to forecast, using `simulate_infections()` from EpiNow2.
We set a constant reproduction number above 1 and a fixed population size, which produces an epidemic that grows, peaks, and then declines as susceptibles are depleted.
Setting `pop_period = "all"` applies the depletion across the whole simulated period, so the turnover is built into the data-generating process.
The generation time and delays are the example distributions supplied with EpiNow2, fixed to point values with `fix_parameters()` because simulation requires known parameters.

```{r simulate}
gen_time <- fix_parameters(example_generation_time)
delays <- fix_parameters(example_incubation_period + example_reporting_delay)
pop_size <- 100000
dates <- seq.Date(as.Date("2023-01-01"), by = "day", length.out = 170)
R <- data.frame(date = dates, R = 1.25)

sim <- simulate_infections(
  R = R,
  initial_infections = 20,
  generation_time = gt_opts(gen_time),
  delays = delay_opts(delays),
  obs = obs_opts(family = "poisson"),
  pop = Fixed(pop_size),
  pop_period = "all"
)

reported <- sim |>
  as_tibble() |>
  filter(variable == "reported_cases") |>
  transmute(date, confirm = value)
```

The simulated series of reported cases rises to a peak in late March before declining.

```{r plot-outbreak}
ggplot(reported, aes(x = date, y = confirm)) +
  geom_col(fill = "#4292C6") +
  labs(x = "Date", y = "Reported cases", title = "Simulated outbreak") +
  theme_bw()
```

The decline happens because susceptibles are running out — the population at risk is finite, and once enough people have been infected, the effective reproduction number drops below 1 even though the underlying $R_t$ is still 1.25.
This is the mechanism we want our forecasts to capture.

# A single forecast

To forecast we truncate the series at a forecast date and fit only to the data available up to that point.
We choose a date shortly before the peak, where the epidemic is still growing and the turnover has not yet been observed.

```{r truncate}
forecast_date <- as.Date("2023-03-18")
train <- reported |> filter(date <= forecast_date)
```

We fit two models: the EpiNow2 default renewal model and the ARIMA baseline from earlier sessions.
Both use a 14-day forecast horizon.

```{r single-forecast, warning = FALSE, message = FALSE}
en2_default <- estimate_infections(
  train,
  generation_time = gt_opts(example_generation_time),
  delays = delay_opts(example_incubation_period + example_reporting_delay),
  rt = rt_opts(rw = 7),
  gp = NULL,
  forecast = forecast_opts(horizon = 14),
  stan = stan_config
)

arima_default <- train |>
  as_tsibble(index = date) |>
  model(ar2 = ARIMA(my_fourth_root(confirm)))
```

We plot both forecasts against the data that were later observed.
The shaded ribbons are the 50% and 90% credible intervals, the lines are the posterior medians, and the points are the observations split into the data used for fitting and the held-out truth.

```{r plot-single-forecast}
en2_fc_sum <- summary(en2_default, type = "parameters") |>
  as_tibble() |>
  filter(variable == "reported_cases", type == "forecast") |>
  transmute(date, model = "epi_default", median, lower_50, lower_90, upper_50, upper_90)

arima_fc_sum <- generate(arima_default, h = 14, times = 1000) |>
  as_tibble() |>
  group_by(date) |>
  summarise(
    median = median(.sim),
    lower_50 = quantile(.sim, 0.25),
    lower_90 = quantile(.sim, 0.05),
    upper_50 = quantile(.sim, 0.75),
    upper_90 = quantile(.sim, 0.95),
    .groups = "drop"
  ) |>
  mutate(model = "arima")

single_fc_sum <- bind_rows(en2_fc_sum, arima_fc_sum)

ggplot(single_fc_sum, aes(x = date, fill = model, colour = model)) +
  geom_ribbon(aes(ymin = lower_90, ymax = upper_90), alpha = 0.15) +
  geom_ribbon(aes(ymin = lower_50, ymax = upper_50), alpha = 0.3) +
  geom_line(aes(y = median)) +
  geom_point(
    data = reported |> filter(date > forecast_date - 21, date <= forecast_date + 14),
    aes(x = date, y = confirm),
    inherit.aes = FALSE
  ) +
  geom_vline(xintercept = forecast_date, linetype = "dashed") +
  scale_fill_manual(values = c("epi_default" = "#4292C6", "arima" = "#E6550D")) +
  scale_colour_manual(values = c("epi_default" = "#4292C6", "arima" = "#E6550D")) +
  labs(
    x = "Date", y = "Reported cases",
    title = "14-day forecast from just before the peak"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

Neither model knows that the epidemic is about to turn over.
The ARIMA projects the recent trend forward; the renewal model does the same, because the default model has no mechanism for susceptible depletion.
Both over-shoot the peak — foreshadowing the backtest, where we give one model the ability to see the turnover coming.

::: callout-tip
## Take 5 minutes

Look at the two forecast ribbons against the held-out observations.
Which model looks closer to the truth over the first few days, and which drifts further off as the horizon grows?
What single piece of information would either model need to predict the turning point?
:::

# The models we will compare

We compare four models that differ in how much epidemiological mechanism they include.

- **ARIMA** — `ARIMA(my_fourth_root(confirm))` with auto-ARIMA, so fable selects the order automatically from the data. Auto-ARIMA may select different orders at different forecast dates — this is a feature of a fair statistical baseline, not a bug.
- **default** — renewal + delays, no depletion. Projects recent growth through the peak, so we expect it to over-shoot.
- **fixed_pop** — adds depletion with the *true* population size. Should turn over at the right time.
- **wrong_pop** — depletion with a population 10× too small. Depletion bites early, so we expect premature turnover and under-shoot. This is the "wrong mechanism" demo.

The EpiNow2 models share the same renewal structure; they differ only in the `pop` argument to `rt_opts()`.
The ARIMA baseline is fit separately with `fable`, using the same fourth-root transformation as in earlier sessions.

```{r model-specs}
epi_models <- list(
  default   = rt_opts(rw = 7),
  fixed_pop = rt_opts(rw = 7, pop = Fixed(pop_size),       future = "latest"),
  wrong_pop = rt_opts(rw = 7, pop = Fixed(pop_size / 10), future = "latest")
)
```

# Backtesting across forecast dates

We backtest each model from 10 forecast dates, one week apart, spanning the growth phase, the peak, and the decline.
Contrasting phases matters because susceptible depletion only shapes the forecast near and after the peak; early in the growth phase it is almost irrelevant.

```{r forecast-dates}
forecast_dates <- seq.Date(
  from = as.Date("2023-02-15"), to = as.Date("2023-04-19"), by = "week"
)
length(forecast_dates)
```

We define a helper that fits an EpiNow2 model with variational Bayes and returns sample-level predictions, and another that does the same for the ARIMA baseline via `fable::generate()`.

```{r fit-helpers}
fit_epinow2 <- function(rt, data) {
  estimate_infections(
    data,
    generation_time = gt_opts(example_generation_time),
    delays = delay_opts(example_incubation_period + example_reporting_delay),
    rt = rt, gp = NULL,
    forecast = forecast_opts(horizon = 14),
    stan = stan_config
  )
}

forecast_arima_samples <- function(train, horizon = 14, times = 1000) {
  fit <- train |>
    as_tsibble(index = date) |>
    model(ar2 = ARIMA(my_fourth_root(confirm)))
  generate(fit, h = horizon, times = times) |>
    as_tibble() |>
    transmute(
      forecast_date = max(train$date),
      date,
      horizon = as.integer(date - max(train$date)),
      sample = as.integer(.rep),
      predicted = .sim,
      model = "arima"
    )
}
```

We loop over the forecast dates, fitting all four models at each date and combining the sample-level predictions into a single tibble.

```{r backtest, results = "hide", warning = FALSE, message = FALSE}
backtest_one <- function(fdate) {
  train <- reported |> filter(date <= fdate)
  en2 <- map_dfr(names(epi_models), function(m) {
    fit <- fit_epinow2(epi_models[[m]], train)
    get_predictions(fit, format = "sample") |>
      as_tibble() |>
      transmute(
        forecast_date = fdate,
        date, horizon, sample, predicted, model = m
      )
  })
  arm <- forecast_arima_samples(train)
  bind_rows(en2, arm)
}

all_forecasts <- map_dfr(forecast_dates, backtest_one) |>
  filter(!is.na(predicted))
```

## Visualising the backtested forecasts

Before scoring, we look at the forecast trajectories themselves.
Each panel is a forecast date; the ribbons are the 50% and 90% credible intervals, the lines are the posterior medians, and the points are the observed data.

```{r plot-backtests, fig.height = 10}
trajectory_summary <- all_forecasts |>
  filter(horizon > 0) |>
  group_by(model, forecast_date, date) |>
  summarise(
    median = median(predicted),
    lower_50 = quantile(predicted, 0.25),
    lower_90 = quantile(predicted, 0.05),
    upper_50 = quantile(predicted, 0.75),
    upper_90 = quantile(predicted, 0.95),
    .groups = "drop"
  )

obs_windows <- map_dfr(forecast_dates, function(fd) {
  reported |>
    filter(date >= fd - 28, date <= fd + 14) |>
    mutate(forecast_date = fd)
})

ggplot(trajectory_summary, aes(x = date)) +
  geom_ribbon(aes(ymin = lower_90, ymax = upper_90, fill = model), alpha = 0.15) +
  geom_ribbon(aes(ymin = lower_50, ymax = upper_50, fill = model), alpha = 0.3) +
  geom_line(aes(y = median, colour = model)) +
  geom_point(data = obs_windows, aes(y = confirm), size = 0.5) +
  geom_vline(
    data = tibble(forecast_date = forecast_dates),
    aes(xintercept = forecast_date),
    linetype = "dashed"
  ) +
  facet_wrap(~forecast_date, scales = "free", ncol = 1) +
  labs(
    x = "Date", y = "Reported cases", colour = "Model", fill = "Model",
    title = "Backtested forecasts by forecast date"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

During growth the models agree and all track the data.
Around the peak the default model keeps climbing while the fixed-population model bends over with the observations; the wrong-population model turns over too early.
In the decline the recent downturn is already visible in the data, so all models project a fall and the gap between them narrows again.

::: callout-tip
## Take 5 minutes

Before reading on, predict which model you expect to score best overall, and at which forecast dates you expect the models to differ most.
Does your prediction change if we score on the natural scale versus the log scale?
:::

::: {.callout-note collapse="true"}
## Solution

The fixed-population model should score best overall, because it is the only one that both knows the turnover is coming and times it correctly.
The models should differ most around the peak, where depletion is decisive; during growth and decline the data already contain the trend, so the models converge.
:::

# Scoring

We merge the forecasts with the observed values and build a single `forecast_sample` object, keeping only out-of-sample horizons (`horizon > 0`).

```{r build-forecast-sample}
obs <- reported |> transmute(date, observed = confirm)

score_data <- all_forecasts |>
  filter(horizon > 0) |>
  left_join(obs, by = "date")

fc <- as_forecast_sample(
  score_data,
  forecast_unit = c("model", "forecast_date", "date", "horizon"),
  observed = "observed",
  predicted = "predicted",
  sample_id = "sample"
)
```

Reported cases span several orders of magnitude over an epidemic, so a score on the natural scale is dominated by the largest counts near the peak.
As in the [forecast evaluation session](forecast-evaluation.qmd), we score on both the natural and the log scale; we use `transform_forecasts()` with `log_shift` to add log-transformed forecasts alongside the originals, which `score()` then scores together, labelling each with a `scale` column.

```{r score}
fc <- transform_forecasts(fc, fun = log_shift, offset = 1)
scores <- score(fc)
```

## Overall scores by model

```{r scores-by-model}
overall <- summarise_scores(scores, by = c("model", "scale")) |>
  as_tibble() |>
  mutate(across(where(is.numeric), \(x) signif(x, 3))) |>
  arrange(scale, crps)

overall |>
  select(model, scale, crps, bias, dispersion, underprediction, overprediction)
```

On the natural scale the fixed-population model scores best on CRPS, even though its bias is not the smallest.
On the log scale the ARIMA model edges ahead, because log scoring down-weights the large counts near the peak where the epidemiological models excel.
CRPS rewards being close and well-calibrated across the whole predictive distribution, not just unbiased on average.
The default model has a large positive bias: it over-predicts, because it projects growth through the peak.
The wrong-population model under-predicts, because depletion bites too early — its underprediction component is high, though its average bias is near zero because the errors balance across dates.
The ARIMA model has a low average bias — but only because its errors point in different directions on different dates, which its mid-pack natural-scale CRPS exposes.
As we discussed in the [session on evaluating forecasts](forecast-evaluation.qmd), bias and CRPS measure different aspects of a probabilistic forecast, so a near-zero bias can mask a model that is wrong in both directions.

::: callout-tip
## Take 5 minutes

On the natural scale, the fixed-population model wins on CRPS, but the ARIMA has a smaller average bias.
Recall from the [forecast evaluation session](forecast-evaluation.qmd) what CRPS and bias each measure: which metric should you trust when they disagree, and why?
:::

## Scores by forecast date

Breaking the natural-scale CRPS down by forecast date shows how the ranking shifts across the epidemic's phases.

```{r scores-by-date}
by_date <- summarise_scores(
  scores, by = c("model", "forecast_date", "scale")
) |>
  as_tibble()

by_date |>
  filter(scale == "natural") |>
  arrange(forecast_date, crps) |>
  select(forecast_date, model, crps) |>
  mutate(crps = round(crps, 1))
```

We can also visualise this as a plot, with the observed case curve below for context.

```{r plot-scores-by-date}
scores_plot <- by_date |>
  filter(scale == "natural") |>
  ggplot(aes(x = forecast_date, y = crps, colour = model)) +
  geom_point() +
  geom_line() +
  labs(x = "Forecast date", y = "CRPS (natural scale)", colour = "Model") +
  theme_bw() +
  theme(legend.position = "bottom")

obs_plot <- reported |>
  filter(date >= min(forecast_dates), date <= max(forecast_dates) + 14) |>
  ggplot(aes(x = date, y = confirm)) +
  geom_col(fill = "#4292C6") +
  labs(x = "Date", y = "Reported cases") +
  theme_bw()

gridExtra::grid.arrange(scores_plot, obs_plot)
```

During growth (early dates) the models are close.
Around the peak (late March) the fixed-population model pulls well ahead as the default over-projects the turnover.
In the decline (April) the models converge again, because the downturn is already in the data by the time the forecast is made.

## Scores by horizon

```{r scores-by-horizon}
by_horizon <- summarise_scores(
  scores, by = c("model", "horizon", "scale")
) |>
  as_tibble()

ggplot(by_horizon, aes(x = horizon, y = crps, colour = model)) +
  geom_line() +
  geom_point(size = 1) +
  facet_wrap(~scale, scales = "free_y") +
  labs(
    x = "Forecast horizon (days)", y = "Mean CRPS", colour = "Model",
    title = "Forecast error by horizon and scale"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

On the natural scale the gap between models opens at longer horizons, where the forecasts run into the peak.
On the log scale the differences are smaller and steadier, because log scoring down-weights the large counts that dominate the natural scale.

## Coverage

Interval coverage is not returned by `score()` for sample forecasts, so we compute it manually: for each forecast, we check whether the observation falls within the 50% and 90% prediction intervals, then average across all forecasts for each model.

```{r coverage}
coverage_data <- score_data |>
  group_by(model, forecast_date, date, horizon) |>
  summarise(
    in_50 = observed[1] >= quantile(predicted, 0.25) &
            observed[1] <= quantile(predicted, 0.75),
    in_90 = observed[1] >= quantile(predicted, 0.05) &
            observed[1] <= quantile(predicted, 0.95),
    .groups = "drop"
  ) |>
  group_by(model) |>
  summarise(
    interval_coverage_50 = mean(in_50),
    interval_coverage_90 = mean(in_90)
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  arrange(model)

coverage_data
```

Well-calibrated models should hit approximately 50% and 90% coverage.
Variational Bayes gives approximate posteriors, so coverage may deviate from nominal levels; the relative ranking between models is more informative than the absolute values.

# What goes wrong with the wrong mechanism

To see the cost of a wrong mechanism clearly, we isolate the `wrong_pop` and `fixed_pop` models at a forecast date just before the peak.
These two models are identical except for one number: the population size.

```{r wrong-mechanism}
peak_date <- as.Date("2023-03-15")

mechanism_compare <- all_forecasts |>
  filter(forecast_date == peak_date, model %in% c("fixed_pop", "wrong_pop"), horizon > 0) |>
  group_by(model, date) |>
  summarise(
    median = median(predicted),
    lower_90 = quantile(predicted, 0.05),
    upper_90 = quantile(predicted, 0.95),
    .groups = "drop"
  )

ggplot(mechanism_compare, aes(x = date, fill = model, colour = model)) +
  geom_ribbon(aes(ymin = lower_90, ymax = upper_90), alpha = 0.2) +
  geom_line(aes(y = median)) +
  geom_point(
    data = reported |> filter(date > peak_date - 14, date <= peak_date + 14),
    aes(x = date, y = confirm),
    inherit.aes = FALSE
  ) +
  geom_vline(xintercept = peak_date, linetype = "dashed") +
  labs(
    x = "Date", y = "Reported cases",
    title = "Right vs wrong population size, forecast from 2023-03-15"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

The `fixed_pop` model (correct population) bends over with the observations.
The `wrong_pop` model (population 10× too small) turns over far too early — it predicts the epidemic is already ending when it is still growing.
The only difference between these two models is one number, and it moves the turning point by weeks.
This is the cost of a wrong mechanism: not noise, but a *systematic* error in a predictable direction.

# Going further

## Challenge

-   Re-run the backtest with an *uncertain* population, `pop = Normal(mean = pop_size, sd = pop_size / 2)`. Does estimating the population recover the fixed-pop performance, or does the data fail to identify it?
-   Add a fifth EpiNow2 model with a Gaussian process (`gp = gp_opts()`, from EpiNow2) instead of the random walk. Does the smoother $R_t$ change the peak forecast?
-   Extend the forecast horizon to 28 days. Which model degrades fastest?

## Methods in practice

-   @abbottEpiNow2EstimateRealtime2025 — EpiNow2.
-   @bhattSemiMechanisticBayesianModeling2020 — renewal models in practice.
-   @gosticPracticalConsiderationsMeasuring2020 — practical $R_t$.

# Wrap up

-   Review what you've learned in this session with the [learning objectives](../reference/learning_objectives.qmd)
-   Share your [questions and thoughts](../reference/help.qmd)

## References

::: {#refs}
:::
