---
title: "Evaluating real-world ensembles"
order: 8
bibliography: ../nfidd.bib
---

# Introduction

This session explicitly synthesizes content from two previous sessions, as we will be [building and evaluating ensembles](forecast-ensembles.qmd) using [real-world forecast data](real-world-forecasts.qmd). In particular, we will explore weighting models in the ensemble.

## Slides

-   [Real-world ensembles](slides/real-world-ensembles.qmd)

## Objectives

The aim of this session is to develop advanced skills in working with ensemble forecasts and an appreciation for the challenges improving on equal-weight ensembles by estimating model weights.

::: {.callout-note collapse="true"}
# Setup

## Source file

The source file of this session is located at `sessions/eval-real-world-ensembles.qmd`.

## Libraries used

In this session we will use the `nfidd` package to access some stored datasets, the `dplyr` package for data wrangling, the `ggplot2` library for plotting, and the following [hubverse](https://hubverse.io/) packages: `hubData`, `hubUtils`, `hubEvals` and `hubEnsembles`.

```{r libraries, message = FALSE}
library("nfidd.forecasting")
library("dplyr")
library("ggplot2")
library("hubData")
library("hubUtils")
library("hubEvals")
library("hubEnsembles")
theme_set(theme_bw())
```

::: callout-tip
The best way to interact with the material is via the [Visual Editor](https://docs.posit.co/ide/user/ide/guide/documents/visual-editor.html) of RStudio.
:::

## Initialisation

We set a random seed for reproducibility.
Setting this ensures that you should get exactly the same results on your computer as we do.
This is not strictly necessary but will help us talk about the models.

```{r}
set.seed(78) # for Greg Maddux
```
:::


# Downloading the forecast data

We will start by accessing forecast data from the COVID-19 Forecast Hub as we did in [the earlier session](real-world-forecasts.qmd).

## Accessing forecast data from the cloud

The following code can be used to retrieve COVID-19 forecasts from the cloud.
To make these results more reproducible and to provide a snapshot of the data, we also provide the forecasts as a data object that can be loaded directly from the `nfidd` package, although the following code could be run or adapted for filtering different subsets of data (see "Going Further" section.

```{r load-forecasts, eval=FALSE}
hub_path_cloud <- hubData::s3_bucket("s3://covid19-forecast-hub")
hub_con_cloud <- hubData::connect_hub(hub_path_cloud, skip_checks = TRUE)

data(covid_locations)

covid_forecasts <- hub_con_cloud |> 
  filter(
    reference_date <= as.Date("2025-07-12"),
    output_type == "quantile",
    target == "wk inc covid hosp",
    horizon >= 0
    ) |> 
  collect() |> 
  left_join(covid_locations)
```

This line of code will load all the forecasts saved in the package into your R session.

```{r load-saved-forecasts}
data(covid_forecasts)
```

## Accessing target data from the cloud

Here is code to query the time-series data and oracle output data from the cloud-based hub.

```{r load-ts-target-data, eval = FALSE}
covid_time_series <- connect_target_timeseries(hub_path_cloud) |> 
  filter(target == "wk inc covid hosp") |> 
  collect() |> 
  left_join(covid_locations)
```

```{r build-oracle-output-data}
hub_path_cloud <- hubData::s3_bucket("s3://covid19-forecast-hub")
covid_oracle_output <- connect_target_oracle_output(hub_path_cloud) |> 
  collect()
covid_oracle_output
```

However, to ensure reproducibility (since this is a live dataset), we have downloaded this object for you already as of July 9, 2025, and it is available to load in from the course R package directly.

```{r load-ts-target-data-archived}
data(covid_time_series)
```

# Ensembling forecasts

Recall that in the earlier session we analyzed and compared individual forecasts from the COVID-19 forecast hub, including the `CovidHub-ensemble` model put together by the hub administrators. 
In this session, we will build several new ensemble forecasts and compare these to the `CovidHub-ensemble` model.
Based on the last results from the section above, we saw that there was some fairly consistent performance from a few of the models.
This is one precondition for weighted ensembles to have a chance at performing well.

We will use ensembling methods and tools presented in the [Forecast Ensembles session](forecast-ensembles.qmd).
Here are the different approaches we will use:

-   `lop_unweighted_all`: an unweighted linear opinion pool (LOP) of all available forecasts each week.
-   `lop_unweighted_select`: an unweighted LOP of the most commonly submitted forecasts.
-   `lop_inv_wis_weighted_select`: a weighted LOP of models that submitted more than 60% of the time.
-   `median_inv_wis_weighted_select`: a weighted median ensemble of models that submitted more than 60% of the time.

Note that the existing `CovidHub-ensemble` forecast is basically `median-unweighted-all` in the above nomenclature.
It would be hard to do any `weighted-all` ensembles since when models are missing a lot of forecasts it might be hard to estimate an accurate weight for them.

## Finding frequently submitted models
Let's start by identifying the subset of models that submitted more than 60% of the time.
```{r}
## 33 dates, 4 horizons, 53 locations, 60%
threshold_60pct <- 33 * 4 * 53 * 0.6
model_subset <- covid_forecasts |> 
  filter(output_type_id == 0.5) |> 
  group_by(model_id) |> 
  summarize(targets = n()) |> 
  filter(targets > threshold_60pct) |> 
  pull(model_id)
model_subset
```


## Build unweighted LOP ensembles

Let's start with the "easy" part, building the unweighted ensembles.

::: {.callout-tip collapse="true"}
### LOP details

Note that when we build LOPs with quantile based forecasts, the algorithm needs to approximate a full distribution (the tails are not well-defined in a quantile representation).
There are some assumptions made under the hood about how this is approximated.
For the really gory details, you can check out [the `hubEnsembles` documentation](https://hubverse-org.github.io/hubEnsembles/dev/articles/hubEnsembles.html?q=weight#creating-ensembles-with-linear_pool)
:::

Note that trying to build all of the forecasts in one call to `linear_pool()` exceeded the memory on your instructors laptop, so we will do a `purrr::map_dfr()` "loop" to do this one reference_date at a time.
(This code takes a few minutes to run.)

```{r build-unweighted-ensembles}
# Get a list of unique reference dates, removing the last one
reference_dates <- covid_forecasts |> 
  filter(reference_date != as.Date("2025-07-12")) |> 
  pull(reference_date) |> 
  unique()

# Map over each date and apply linear_pool separately
lop_unweighted_all <- purrr::map_dfr(
  reference_dates, 
  function(date) {
    covid_forecasts |> 
      filter(model_id != "CovidHub-ensemble", 
             reference_date == date) |> 
      hubEnsembles::linear_pool(model_id = "lop_unweighted_all")
})


lop_unweighted_select <- purrr::map_dfr(
  reference_dates, 
  function(date) {
    covid_forecasts |> 
      filter(model_id %in% model_subset,
             model_id != "CovidHub-ensemble", 
             reference_date == date) |> 
      hubEnsembles::linear_pool(model_id = "lop_unweighted_select")
})
```

## Build weighted ensemble

Inverse WIS weighting is a simple method that weights the forecasts by the inverse of their WIS over some period.
Note that identifying what this period should be in order to produce the best forecasts is not straightforward as predictive performance may vary over time if, say, some models are better at making predictions during different parts of an outbreak.
For example, some models might be good at predicting exponential growth at the beginning of an outbreak but not as good at seeing the slowdown.

The main benefit of WIS weighting over other methods is that it is simple to understand and implement.
However, it does not optimise the weights directly to produce the best forecasts.
It relies on the hope that giving more weight to better performing models yields a better ensemble.

We will use the WIS scores that we have computed above to generate weights for our ensembles.
The weights for one `reference_date` will be based on the inverse of the cumulative average WIS achieved by the model up to the week prior to the current `reference_date`.

::: callout-caution
Note that since our scores are computed based on data available at the end of the season, this is kind of "cheating", since we're using information that would not have been available to us in real time.
If we were to do this entirely correctly, we would, for each week, compute the scores based on the available data at that time.
But, for instructional purposes, this will suffice.
:::

Let's start by calculating WIS values by model and `target_end_date`.
This will allow us to see what scores are available for a model's forecasts that it had for dates prior to the current forecast date.

```{r scores-by-target-end-date}
scores_by_target_date <- covid_forecasts |> 
  filter(model_id %in% model_subset,
         location != "US") |> 
  score_model_out(
    covid_oracle_output,
    metrics = "wis",
    by = c("model_id", "target_end_date")
  )
scores_by_target_date
```

Now, we will compute a cumulative average WIS up to a given date for all models except the ensemble, which we are not going to include in our ensemble.

```{r cumulate-wis}
cum_wis <- scores_by_target_date |>
  filter(model_id != "CovidHub-ensemble") |> 
  arrange(model_id, target_end_date) |> 
  group_by(model_id) |> 
  mutate(
    cumulative_mean_wis = cummean(wis), ## cumulative mean WIS including current week
    cumulative_mean_wis_lag1 = lag(cumulative_mean_wis)
    ) |> 
  ungroup()
cum_wis
```

::: callout-tip
Note that we also compute a `cumulative_mean_wis_lag1`, which is the mean up through the previous week.
This is to inject a bit of realism: how could we know in real-time how our forecast did for the current week?

So the `cumulative_mean_wis_lag1` is the cumulative mean WIS that we could have observed at the `target_end_date`.
(This all still assumes that whatever data we observed at the `target_end_date` was complete and never revised.)
:::

We will proceed by computing the inverse mean WIS and using those to derive the weights.

```{r compute-weights}
weights_per_model <- cum_wis |> 
  mutate(inv_wis = 1 / cumulative_mean_wis_lag1) |>
  group_by(target_end_date) |> 
  mutate(inv_wis_total_by_date = sum(inv_wis, na.rm = TRUE)) |>
  ungroup() |> 
  mutate(weight = inv_wis / inv_wis_total_by_date) |> ## this normalises the weights to sum to 1
  select(model_id, target_end_date, weight)
```

And we will put in place a few small fixes to the weights:

-   assign equal weights for the first two weeks. (There are missing forecasts here, and we haven't learned much about the model performance yet anyways.)
-   change `target_end_date` to `reference_date` because now we want the weights computed up to that `target_end_date` to be applied on that `reference_date`.
-   set to 0 a missing weight for the `NEU_ISI-AdaptiveEnsemble`.

```{r fix-weights}
## assign equal weights for the first week
first_week_idx <- weights_per_model$target_end_date == as.Date("2024-11-23")
weights_per_model[first_week_idx, "weight"] <- 1/sum(first_week_idx)

second_week_idx <- weights_per_model$target_end_date == as.Date("2024-11-30")
weights_per_model[second_week_idx, "weight"] <- 1/sum(second_week_idx)

weights_per_model <- rename(weights_per_model, reference_date = target_end_date)

weights_per_model$weight[is.na(weights_per_model$weight)] <- 0
```

As always, it is good to plot your data as a sanity check.
Here are the model weights plotted over time.
For each week (x-axis), the weights assigned to each model are shown in a stacked bar plot, with the height of the bar equal to the weight of the model.

```{r plot-weights}
plot_weights <- ggplot(weights_per_model) +
  geom_col(aes(x = reference_date, y = weight, fill = model_id))
plotly::ggplotly(plot_weights)
```

::: callout-tip
### Take 5 minutes

Looking at the above plot, what is the "story" that the weights tell when plotted over time?
Which models did the ensemble like and dislike over time and how did it change?
(Remember, in the first two weeks, all models received equal weight.)
:::

::: {.callout-note collapse="true"}
### Solution

Starting in the third week, three models were assigned less weight than others: `CMU-climate-baseline`, `CMU-TimeSeries` and `COVIDHub-baseline`.
Over time, `CMU-TimeSeries` and `COVIDHub-baseline` earned back some "trust" with higher weights, but `CMU-climate-baseline` stayed with a low weight.

In general, the weights did not change much after the first few weeks (except in the last week when `UM-DeepOutbreak` wasn't submitted.)
:::

Now that we have the weights and we've checked them, let's start building some ensembles!
First, we will build the median ensembles, both weighted and unweighted.

```{r build-ensemble}
## build the weighted ensemble
median_inv_wis_weighted_select <- covid_forecasts |> 
  filter(model_id %in% model_subset,
         model_id != "CovidHub-ensemble") |> 
  simple_ensemble(weights = weights_per_model, 
                  agg_fun = median,
                  model_id = "median_inv_wis_weighted_select")

## build unweighted select median ensemble
median_unweighted_select <- covid_forecasts |> 
  filter(model_id %in% model_subset,
         model_id != "CovidHub-ensemble") |> 
  simple_ensemble(agg_fun = median,
                  model_id = "median_unweighted_select")
```

And then we will build the inverse WIS weighted LOP ensemble.

::: callout-caution
It takes several minutes for the LOP ensembles to be built if you are running the code on your machine.
:::

```{r lop-weighted}
lop_inv_wis_weighted_select <- purrr::map_dfr(
  reference_dates, 
  function(date) {
    covid_forecasts |> 
      filter(model_id %in% model_subset,
             model_id != "CovidHub-ensemble", 
             reference_date == date) |> 
      linear_pool(model_id = "lop_inv_wis_weighted_select",
                  weights = weights_per_model)
})
```

## Visually compare ensembles

As always, it is really important to **look at the forecasts** to develop some intuition about the similarities and differences between the models. The following code subsets the output to look just at the ensembles (including the original `CovidHub-ensemble`) and then plots forecasts for three locations (GA, MA and CA) at one date. Feel free to modify the code to look at other locations or dates.

```{r assemble-all-forecasts}
covid_forecasts_w_ensembles <- bind_rows(
  covid_forecasts,
  lop_unweighted_all,
  lop_unweighted_select,
  lop_inv_wis_weighted_select,
  median_inv_wis_weighted_select,
  median_unweighted_select
)

model_subset_ens <- c("CovidHub-ensemble", 
                      "lop_unweighted_all",
                      "lop_unweighted_select",
                      "median_inv_wis_weighted_select",
                      "median_unweighted_select",
                      "lop_inv_wis_weighted_select")
```


```{r plot-ensembles}
covid_forecasts_w_ensembles |> 
  filter(
    reference_date %in% as.Date("2025-01-18"),
    abbreviation %in% c("GA", "MA", "CA"),
    model_id %in% model_subset_ens
  ) |> 
  hubVis::plot_step_ahead_model_output(
    target_data = covid_time_series |> 
      filter(as_of == as.Date("2025-03-05"),
             abbreviation %in% c("GA", "MA", "CA")),
    use_median_as_point = TRUE,
    facet = "abbreviation",
    facet_nrow = 3,
    facet_scales = "free_y",
    intervals = 0.95,
    x_col_name = "target_end_date",
    x_target_col_name = "date",
    pal_color = "Set3", 
    title = "Weekly hospitalizations due to COVID-19 (data and forecasts)"
  )
```

::: callout-tip
## Take 5 minutes

Look at the forecasts above, and possibly explore other dates and/or locations. Are there noticeable differences between the point estimates or between the prediction interval widths? If so, what are they?

:::

::: {.callout-note collapse="true"}
## Solution

In the plot shown above, there are not noticeable differences between point estimates. All of the LOP models have wider PIs compared to the other ensembles.

:::


## Compare all ensembles

We will compare the performance of all ensembles just at the state level based on quantitative scoring rules, since the scores by reference date were only computed at the state level.

```{r compare-ensemble-scores}
covid_forecasts_w_ensembles |> 
  filter(model_id %in% model_subset_ens,
         location != "US") |> 
  score_model_out(
    covid_oracle_output
  ) |> 
  arrange(wis) |> 
  knitr::kable(digits = 2)
```

Remember, in the above table, the `CovidHub-ensemble` is kind of like `median-unweighted-all` because it uses a median ensemble of all models available in every week without weights.
Overall, the ensembles have fairly similar average WIS values across all dates and states.
Using a median unweighted ensemble of frequently submitting models marginally outperformed the actual `CovidHub-ensemble`.

::: callout-tip
### Take 5 minutes

Why do you think weighting does not improve the forecasts more?

Seeing these patterns, and the plots of the weights above, is there anything you might change if you were given the opportunity to compute the weights in a different way?

Write down your thoughts about these questions and compare answers with a neighbor.
:::

::: {.callout-note collapse="true"}
### Solution

#### Why aren't weighted ensembles better?

When model performance varies a lot from one week to the next, it may be hard to estimate/anticipate/predict which model should get more weight in a given week.

#### What might your consider changing about the weights?

There isn't one right answer here, but some things that you might consider to do differently are:

1.  Figuring out a way to encourage the model to stay closer to equal weights early on. The weights are being perhaps overly influenced by small sample sizes there.
2.  Thinking about using a rolling window of mean WIS to compute weights rather than overall cumulative. This could be good, but also assumes that recent performance will predict future performance, which is not always true!
:::

It does not seem like weighting adds a lot of value to these forecasts, although perhaps with additional data on model performance the weights could be improved.

We can also look at how the ensemble performance varied over time.

```{r}
ensemble_scores_by_reference_date <- covid_forecasts_w_ensembles |> 
  filter(model_id %in% model_subset_ens,
         location != "US") |> 
  score_model_out(
    covid_oracle_output,
    metrics = "wis",
    by = c("model_id", "reference_date")
  )

p <- ggplot(ensemble_scores_by_reference_date) +
  geom_line(aes(x = reference_date, y = wis, color = model_id))
plotly::ggplotly(p)
```

It is interesting to note how while the LOP ensembles did slightly worse overall, they actually had slightly more accurate forecasts near the peak of the season, especially in the weeks where the scores were the highest. Overall, though, weighting does not seem to have a large impact on the forecast accuracy.


# Going further

## Challenge

- This session only includes data from the year leading up to July 2025. Adjust the code to download data from the past year and re-run analyses on the most recent year of forecasts. Consider adding an ensemble where you manually pick a few models and assign them "human judgment" weights. Compare the results from those in the session above.


## Methods in practice

Here are a set of papers that talk about attempts to build ensembles from a set of models:

- An effort using FluSight models that generated multiple years of prior forecasts for training the weights. [@reich_accuracy_2019]
- Building ensembles to predict dengue in Vietnam.[@colon-gonzalez_probabilistic_2021]
- Predicting health care demand in France during COVID-19. [@paireau_ensemble_2022]
- Predicting COVID-19 in Europe using ensembles. [@sherratt_predictive_2023]
- Evaluating ensembling strategies for COVID-19 forecasts in the US. [@ray_comparing_2023]
- Some work on adaptively estimating weights as functions of other variables. [@ray_prediction_2018;@yamana_superensemble_2016;@mcandrew_adaptively_2021]

# Wrap up

-   Review what you've learned in this session with the [learning objectives](../reference/learning_objectives.qmd)

# References

::: {#refs}
:::
