---
title: "Forecasting with epidemiological insights"
order: 10
bibliography: ../nfidd.bib
---

```{r echo=FALSE}
options(width = 100)
```

# Introduction

The statistical models so far (ARIMA, seasonal baselines) know nothing about disease transmission.
Here we build a **semi-mechanistic** forecast: estimate the time-varying reproduction number $R_t$ from the renewal equation, then simulate incidence forward — letting $R_t$ take a random walk and adding negative-binomial observation noise.
We wrap it as a **`fable` model**, so students use it exactly like `NAIVE()` or `ARIMA()`.
This builds on the renewal-equation material in the companion [nowcasting course](https://nfidd.github.io/sismid-nowcasting/).

::: callout-caution
Prototype session — minimal prose. The `RENEWAL()` model is defined inline here; in the real course it would move to the package's `R/` and be exported.
:::

::: {.callout-note collapse="true"}
## Setup

```{r libraries, message = FALSE}
library("dplyr")
library("ggplot2")
library("epidatr")
library("tsibble")
library("fabletools")
library("fable")          # NAIVE() and statistical baselines
library("distributional")
library("EpiEstim")
library("ggtime")
## NB: MASS is used via MASS::glm.nb() (not attached — it masks dplyr::select)
set.seed(123)
```

### The `RENEWAL()` model (prototype)

```{r renewal-model}
## one simulated renewal trajectory of length h from a fitted RENEWAL object
.renewal_path <- function(object, h) {
  w <- object$gi[-1]; K <- length(w)
  r0 <- EpiEstim::sample_posterior_R(object$rt, 1)
  rtj <- exp(log(r0) + cumsum(rnorm(h, 0, object$sigma_rw)))   # Rt random walk
  series <- object$y; out <- numeric(h)
  for (k in seq_len(h)) {
    foi <- sum(rev(utils::tail(series, K)) * w)                # renewal force of infection
    out[k] <- rnbinom(1, size = object$theta, mu = rtj[k] * foi)  # negbin observation
    series <- c(series, out[k])
  }
  out
}

train_renewal <- function(.data, specials, gi = c(0, 0.65, 0.30, 0.05),
                          window = 4L, sigma_rw = 0.1, n_sim = 2000, ...) {
  mv <- tsibble::measured_vars(.data); y <- .data[[mv]]; n <- length(y)
  t_start <- 2:(n - window + 1); t_end <- t_start + (window - 1)
  rt <- suppressMessages(EpiEstim::estimate_R(
    y, method = "non_parametric_si",
    config = EpiEstim::make_config(list(si_distr = gi, t_start = t_start, t_end = t_end))))
  dat <- data.frame(y = y, t = seq_len(n))
  theta <- tryCatch(suppressWarnings(MASS::glm.nb(y ~ splines::ns(t, df = 10), data = dat)$theta),
                    error = function(e) 50)
  structure(list(y = y, gi = gi, rt = rt, theta = theta,
                 sigma_rw = sigma_rw, n_sim = n_sim),
            class = "RENEWAL")
}

RENEWAL <- function(formula, ...) {
  md <- new_model_class("RENEWAL", train = train_renewal, specials = new_specials())
  new_model_definition(md, !!rlang::enquo(formula), ...)
}

forecast.RENEWAL <- function(object, new_data, specials = NULL, ...) {
  h <- nrow(new_data)
  sims <- replicate(object$n_sim, .renewal_path(object, h))         # h x n_sim
  distributional::dist_sample(lapply(seq_len(h), function(k) sims[k, ]))
}

generate.RENEWAL <- function(x, new_data, specials = NULL, ...) {
  new_data |>
    dplyr::group_by(.data$.rep) |>
    dplyr::mutate(.sim = .renewal_path(x, dplyr::n())) |>
    dplyr::ungroup()
}

## one-step-ahead fitted mean = Rhat_t * (force of infection)_t
fitted.RENEWAL <- function(object, ...) {
  y <- object$y; n <- length(y); w <- object$gi[-1]; K <- length(w)
  lambda <- rep(NA_real_, n)
  for (t in seq_len(n)) if (t > K) lambda[t] <- sum(rev(y[(t - K):(t - 1)]) * w)
  out <- rep(NA_real_, n); R <- object$rt$R
  for (i in seq_along(R$t_end)) {
    tt <- R$t_end[i]
    if (!is.na(lambda[tt])) out[tt] <- R$`Median(R)`[i] * lambda[tt]
  }
  out
}

residuals.RENEWAL <- function(object, ...) object$y - fitted.RENEWAL(object)

model_sum.RENEWAL <- function(x) sprintf("RENEWAL(theta=%.0f)", x$theta)
```
:::

# Data: ILINet counts

Mechanistic models need **case counts** (not the weighted-ILI percentage): the Poisson/negbin renewal machinery only makes sense for integer event counts.
We use ILINet `num_ili` (ILI patient visits).

```{r load-data}
raw <- pub_fluview(regions = "nat", epiweeks = epirange(201335, 201820))
flu_all <- raw |> transmute(region, epiweek, cases = num_ili) |>
  as_tsibble(index = epiweek, key = region)

origin  <- as.Date("2017-12-30")
horizon <- 4L
train   <- flu_all |> filter(epiweek <= origin)
```

# Fit and forecast — like any `fable` model

```{r fit}
fit <- train |>
  model(persistence = NAIVE(cases), renewal = RENEWAL(cases))
fit

fc <- fit |> forecast(h = horizon)
fc |> hilo(90)
```

```{r forecast-plot}
obs_window <- flu_all |> filter(epiweek >= origin - 7 * 20, epiweek <= origin + 7 * horizon)
fc |> autoplot(obs_window) +
  facet_wrap(~.model) +
  labs(title = "Semi-mechanistic RENEWAL vs. a persistence baseline",
       subtitle = "line continues into the forecast period = what actually happened",
       y = "ILI patient visits", x = NULL) +
  theme_bw()
```

# Estimated vs. projected $R_t$

The forecast works by carrying $R_t$ forward as a random walk. We can show both the $R_t$ the model **estimated** from the data and the $R_t$ it **projects** into the future.

```{r rt-plot}
m <- (fit |> pull(renewal))[[1]]$fit          # recover the fitted RENEWAL object

estimated <- m$rt$R |>
  transmute(epiweek = train$epiweek[t_end], kind = "estimated",
            median = `Median(R)`, lower = `Quantile.0.025(R)`, upper = `Quantile.0.975(R)`)

r0   <- EpiEstim::sample_posterior_R(m$rt, 2000)
Rmat <- sapply(seq_along(r0), \(s) exp(log(r0[s]) + cumsum(rnorm(horizon, 0, m$sigma_rw))))
projected <- tibble(
  epiweek = max(train$epiweek) + 7 * seq_len(horizon), kind = "projected",
  median = apply(Rmat, 1, median),
  lower  = apply(Rmat, 1, quantile, 0.025),
  upper  = apply(Rmat, 1, quantile, 0.975))

bind_rows(estimated, projected) |>
  filter(epiweek >= origin - 7 * 20) |>
  ggplot(aes(epiweek, median, colour = kind, fill = kind)) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, colour = NA) +
  geom_line() +
  labs(title = "Rt: estimated (from data) vs. projected (random walk)",
       y = expression(R[t]), x = NULL, colour = NULL, fill = NULL) +
  theme_bw()
```

# Sample trajectories and accuracy

Because it's a generative model, `generate()` draws whole future trajectories (this is what the ensemble/hub sessions consume), and `fitted()`/`residuals()` make `accuracy()` work.

```{r generate}
sims <- fit |> select(renewal) |> generate(h = horizon, times = 50)

ggplot() +
  geom_line(data = obs_window |> filter(epiweek <= origin), aes(epiweek, cases)) +
  geom_line(data = sims, aes(epiweek, .sim, group = .rep), colour = "steelblue", alpha = 0.2) +
  labs(title = "50 simulated renewal trajectories", y = "ILI patient visits", x = NULL) +
  theme_bw()

accuracy(fit)
```

Tuning knobs pass straight through: `RENEWAL(cases, sigma_rw = 0.2)` (faster $R_t$ drift), `sigma_rw = 0` (constant $R_t$), `gi = ...` (generation interval).

# Wrap up

-   Review what you've learned with the [learning objectives](../reference/learning_objectives.qmd)

## References

::: {#refs}
:::
