---
title: "Multivariate forecast evaluation"
order: 9
bibliography: ../nfidd.bib
---

# Introduction

In the last few sessions we built forecasts for many locations and horizons at once and scored them with the CRPS.
We scored each horizon and location on its own and then averaged.
In this session we supplement that "marginal" approach to scoring and comparing models by introducing the energy score.
The energy score is a multivariate extension of the CRPS and allows you to score full trajectories or multiple locations with one score.[@gneiting_assessing_2008;@bay_evaluation_2026;@macarthur_collaborative_2026]

## Slides

-   [Multivariate forecast evaluation](slides/multivariate-forecast-evaluation.qmd)

## Objectives

The aims of this session are to introduce methods for scoring multivariate probabilistic forecasts, for example across time horizons and/or locations.

::: {.callout-note collapse="true"}
# Setup

## Source file

The source file of this session is located at `sessions/multivariate-forecast-evaluation.qmd`.

## Libraries used

In this session we will use the `fable` package for forecasting, the `dplyr`, `tidyr` and `tsibble` packages for data wrangling, the `ggplot2` package for plotting, and the `scoringRules` package for scoring.

```{r libraries, message = FALSE}
library("nfidd.forecasting")
library("dplyr")
library("tsibble")
library("fable")
library("tidyr")
library("ggplot2")
library("scoringRules")
theme_set(theme_bw())
```

::: callout-tip
The best way to interact with the material is via the [Visual Editor](https://docs.posit.co/ide/user/ide/guide/documents/visual-editor.html) of RStudio.
:::

## Initialisation

We set a random seed for reproducibility.

```{r}
set.seed(755) # for Hank Aaron
```

We reuse the same fourth-root transformation and multi-location flu data (`flu_data_hhs`) from the [local hub playground session](hub-playground.qmd).

```{r transform}
fourth_root <- function(x) x^0.25
inv_fourth_root <- function(x) x^4
my_fourth_root <- new_transformation(fourth_root, inv_fourth_root)

data(flu_data_hhs)
```
:::

# From marginal to joint forecasts

A probabilistic time-series forecast that is generated as a realistic trajectory of possible future values is tying together distributions at each timepoint.
Mathematically, the forecast is a joint distribution over the whole *vector* of outcomes, where the vector represents values at time $t+1$ through $t+h$.
When we score each horizon on its own and average, we only ever use the **marginal** forecasts.

This can be important when, for example, you are trying to assess a forecast near a possible peak, and you are trying to assess what the probability is that a surveillance data signal will decline over the coming weeks. 
If all you have is quantiles at each horizon, you can't compute directly what the probability is that the peak will be in a given week.
To make that calculation, you need to see the distribution of the trajectories.

Let's build a **sample-path forecast** from a simple model -- the seasonal AR(2) model from the [forecast evaluation session](forecast-evaluation.qmd) -- for national wILI, and generate many possible *trajectories* for the weeks after our forecast date.

```{r traj-forecast}
h <- 8L
train_cutoff <- as.Date("2018-01-27")   # forecast from around the seasonal peak

## national wILI only, fit on the data available up to the forecast date
us_flu <- flu_data_hhs |> filter(location == "US National")

fit <- us_flu |>
  filter(origin_date <= train_cutoff) |>
  model(fourier_ar2 = ARIMA(my_fourth_root(wili) ~ pdq(2, 0, 0) +
                              fourier(period = "year", K = 3)))

## each replicate (.rep) is one full trajectory over the next h weeks
sims <- fit |>
  generate(h = h, times = 1000, bootstrap = TRUE) |>
  group_by(.rep) |>
  mutate(horizon = row_number()) |>
  ungroup() |>
  as_tibble()
```

Before scoring, let's look at a sample of the forecast trajectories against what was eventually observed.

```{r traj-plot}
## observed trajectory: the actual wILI for the h weeks after the forecast date
observed <- us_flu |> as_tibble() |>
  filter(origin_date > train_cutoff) |>
  arrange(origin_date) |> slice(seq_len(h)) |>
  transmute(horizon = row_number(), wili)

sims |>
  filter(as.integer(.rep) <= 200) |>
  ggplot(aes(horizon, .sim, group = .rep)) +
  geom_line(alpha = 0.1, colour = "#4477AA") +
  geom_line(data = observed, aes(y = wili, group = 1),
            colour = "black", linewidth = 1.1) +
  scale_x_continuous(breaks = seq_len(h)) +
  labs(title = "Trajectory forecast vs. observed (black)",
       x = "horizon (weeks ahead)", y = "wILI %")
```

The forecast is a whole cloud of possible paths. To score the entire path at once, we reshape the samples into a matrix with one row per horizon and one column per sample -- the format the multivariate scoring functions expect (we can't use `fable` for this) -- and pull out the observed vector.

```{r traj-matrix}
samp_wide <- sims |>
  select(horizon, .rep, .sim) |>
  pivot_wider(names_from = .rep, values_from = .sim) |>
  arrange(horizon)

dat_joint <- as.matrix(samp_wide[, -1])   # h rows (horizons) x m columns (samples)
y <- observed$wili                        # the observed trajectory
y
```

# The energy score

The energy score is a proper scoring rule for a **vector** outcome $\mathbf{y}$, and it generalises the CRPS.
Writing $\mathbf{X}, \mathbf{X}'$ for independent draws from the forecast and $\lVert\cdot\rVert$ for Euclidean distance, it balances accuracy against sharpness exactly as the CRPS does, but reads the whole vector at once.
In one dimension it *is* the CRPS.

::: {.callout-tip .optional collapse="true"}
## Mathematical definition (optional)

$$
\text{ES}(F, \mathbf{y}) = \mathbb{E}\lVert \mathbf{X} - \mathbf{y} \rVert - \tfrac{1}{2}\,\mathbb{E}\lVert \mathbf{X} - \mathbf{X}' \rVert
$$

The first term rewards forecasts whose draws land close to the observed vector.
The second term rewards a concentrated cloud of draws.

To compute it, we replace the expectations with averages over our predictive samples.
With $m$ sample trajectories $\mathcal{X}=\{\mathbf{x}_1, \dots, \mathbf{x}_m\}$ from the forecast, we estimate it by

$$
\widehat{\text{ES}}(\mathcal{X},\mathbf{y}) = \frac{1}{m}\sum_{i=1}^{m}\lVert \mathbf{x}_i - \mathbf{y} \rVert - \frac{1}{2m^2}\sum_{i=1}^{m}\sum_{j=1}^{m}\lVert \mathbf{x}_i - \mathbf{x}_j \rVert
$$
:::

::: callout-tip
### Key things to note about the energy score

-   Small values are better, and it is reported in the units of the data.
-   It is a genuine generalisation of the CRPS, to which it reduces in one dimension.
-   It needs **joint samples**, so a forecast that couples the horizons and one that treats them independently are scored differently.
-   Like the CRPS it is best used to **compare** forecasts, not read in isolation.
:::


## Marginal scores don't evaluate the shapes of trajectories

A model that generates trajectories gives us forecasts that have dependence across horizons. But if we scored each horizon on its own and averaged (the marginal CRPS), we would be ignoring the multivariate nature of the forecasts. Using the Energy Score adds a small piece to the score that evaluates how well the trajectories align with the observed trajectory.

To show this in practice, we can build a second "forecast" with __identical marginals__ but without across-horizon dependence: we randomly shuffle the samples independently within each horizon, thus breaking any dependence.
Every horizon keeps exactly the same set of predicted values, so every per-horizon forecast -- and the average CRPS -- is unchanged.

```{r traj-shuffle}
set.seed(1)
## shuffle the samples within each horizon, breaking the links between weeks
dat_shuffled <- t(apply(dat_joint, 1, sample))

## marginal (per-horizon) CRPS, averaged over horizons -- identical for both
crps_joint    <- mean(scoringRules::crps_sample(y = y, dat = dat_joint))
crps_shuffled <- mean(scoringRules::crps_sample(y = y, dat = dat_shuffled))
c(joint = crps_joint, shuffled = crps_shuffled)
```

The energy score, which reads the whole trajectory at once, is not fooled. The "joint" forecast shows a lower (better) energy score, indicating that the original forecast is slightly preferred over the one with independent marginal predictions at each horizon.

```{r traj-es}
es_joint    <- scoringRules::es_sample(y = y, dat = dat_joint)
es_shuffled <- scoringRules::es_sample(y = y, dat = dat_shuffled)
c(joint = es_joint, shuffled = es_shuffled)
```

The plot below shows why. Both panels have the **same marginal forecast at each horizon** -- only the dependence across horizons differs. The original trajectories decline smoothly, like the observed path (black), while the shuffled ones zig-zag across it.

```{r fig-shuffle-cmp, fig.width = 9, fig.height = 3.8}
## reshape each h x m matrix to long format: one line per sample trajectory
to_long <- function(mat, lab) as_tibble(t(mat), .name_repair = ~ paste0("h", seq_along(.x))) |>
  mutate(.rep = row_number()) |> filter(.rep <= 200) |>
  pivot_longer(starts_with("h"), names_to = "horizon", values_to = "wili") |>
  mutate(horizon = as.integer(sub("h", "", horizon)), forecast = lab)

labs <- c(sprintf("Original (ES = %.2f)", es_joint),
          sprintf("Shuffled (ES = %.2f)", es_shuffled))
paths <- bind_rows(to_long(dat_joint, labs[1]), to_long(dat_shuffled, labs[2])) |>
  mutate(forecast = factor(forecast, labs))
obs_both <- expand_grid(forecast = factor(labs, labs), horizon = seq_len(h)) |>
  mutate(wili = rep(y, 2))

ggplot(paths, aes(horizon, wili, group = .rep)) +
  geom_line(alpha = 0.1, colour = "#4477AA") +
  geom_line(data = obs_both, aes(group = 1), colour = "black", linewidth = 1.1) +
  facet_wrap(~ forecast) +
  scale_x_continuous(breaks = seq_len(h)) +
  labs(x = "horizon (weeks ahead)", y = "wILI %")
```

::: callout-tip
## Take 5 minutes

The average CRPS is identical for the two forecasts, but the energy score prefers the original.
Look back at the trajectory plot: why does keeping the links between weeks help here?
:::

::: {.callout-note collapse="true"}
## Solution

The observed data is a **coherent path**: it declines smoothly from the peak, week after week.
The model's own trajectories share some of that coherence, so as whole vectors they stay close to the observed path and trend. (They have some noise, though, so it's not like they are all perfectly aligned trajectories.)
Shuffling keeps each week's marginal forecast intact but breaks the link between weeks, producing jagged trajectories that wander away from the observed path.
The energy score sees this and scores the shuffled forecast worse; the marginal CRPS, which looks at one horizon at a time, cannot.
:::

# Summary

- Using a multivariate score can capture important subtleties about the dependency structure of the forecasts.
- However, often the differences between univariate and multivariate scoring rules are small. The Energy Score still is evaluating distance from the truth, so it might still prefer a tight, jagged forecast that is close to the truth to a wider but smooth forecast that on average isn't as close to the observations.

# Going further

## Challenge

-   There are other dimensions that a model could be joint across, e.g., location. You can fit a genuinely joint-across-locations model with `fable::VAR()` and compare its energy score to an independent ARIMA forecast.
-   Add the variogram score (another multivariate score, available via `scoringRules::vs_sample()`) to the joint comparison. How does it differ from the energy score in what it rewards?

## Methods in practice

-   The `scoringRules` package [documentation](https://cran.r-project.org/package=scoringRules) covers multivariate scores including the energy and variogram scores.
-   Hubs increasingly collect sample (trajectory) forecasts rather than only quantiles, which keeps the joint distribution across locations and horizons.[@hubverseModelOutput2026]
-   @bay_evaluation_2026 describe using the energy score to assess trajectories in longer-term scenario predictions.
-   @macarthur_collaborative_2026 use an adapted energy score to evaluate nowcasts of the proportions of SARS-CoV-2 variants circulating at different timepoints.

# Wrap up

-   Review what you've learned in this session with the [learning objectives](../reference/learning_objectives.qmd)

# References
::: {#refs}
:::
