---
title: "Multivariate forecast evaluation"
order: 9
bibliography: ../nfidd.bib
---

# Introduction

In the last few sessions we built forecasts for many locations and horizons at once and scored them with the CRPS.
We scored each horizon and location on its own and then averaged.
In this session we supplement that "marginal" approach to scoring and comparing models by introducing the energy score.
The energy score is a multivariate extension of the CRPS and allows you to score full trajectories or multiple locations with one score.[@gneiting_assessing_2008;@bay_evaluation_2026;@macarthur_collaborative_2026]

## Slides

-   [Multivariate forecast evaluation](slides/multivariate-forecast-evaluation.qmd)

## Objectives

The aims of this session are to introduce methods for scoring multivariate probabilistic forecasts, for example across time horizons and/or locations.

::: {.callout-note collapse="true"}
# Setup

## Source file

The source file of this session is located at `sessions/multivariate-forecast-evaluation.qmd`.

## Libraries used

In this session we will use the `fable` package for forecasting, the `dplyr`, `tidyr` and `tsibble` packages for data wrangling, the `ggplot2` package for plotting, and the `scoringRules` package for scoring.

```{r libraries, message = FALSE}
library("nfidd.forecasting")
library("dplyr")
library("tsibble")
library("fable")
library("tidyr")
library("ggplot2")
library("scoringRules")
theme_set(theme_bw())
```

::: callout-tip
The best way to interact with the material is via the [Visual Editor](https://docs.posit.co/ide/user/ide/guide/documents/visual-editor.html) of RStudio.
:::

## Initialisation

We set a random seed for reproducibility.

```{r}
set.seed(755) # for Hank Aaron
```

We reuse the same fourth-root transformation and multi-location flu data (`flu_data_hhs`) from the [local hub playground session](hub-playground.qmd).

```{r transform}
fourth_root <- function(x) x^0.25
inv_fourth_root <- function(x) x^4
my_fourth_root <- new_transformation(fourth_root, inv_fourth_root)

data(flu_data_hhs)
```
:::

# From marginal to joint forecasts

A probabilistic time-series forecast that is generated as a realistic trajectory of possible future values is tying together distributions at each timepoint.
Mathematically, the forecast is a joint distribution over the whole *vector* of outcomes, where the vector represents values at time $t+1$ through $t+h$.
When we score each horizon on its own and average, we only ever use the **marginal** forecasts.

This can be important when, for example, you are trying to assess a forecast near a possible peak, and you are trying to assess what the probability is that a surveillance data signal will decline over the coming weeks. 
If all you have is quantiles at each horizon, you can't compute directly what the probability is that the peak will be in a given week.
To make that calculation, you need to see the distribution of the trajectories.

Let's build sample-path forecasts for three regions and pull out the samples we need to score them jointly.

```{r partA-samples}
## fable fits each location separately, so its joint forecast assumes
## independence across regions. A genuinely joint model (e.g. fable::VAR())
## would instead learn the cross-region dependence.
sel_locations <- c("HHS Region 1", "HHS Region 2", "HHS Region 3")
train_cutoff  <- as.Date("2019-02-16")
horizon       <- 4L

fit_joint <- flu_data_hhs |>
  filter(location %in% sel_locations, origin_date <= train_cutoff) |>
  model(arima210 = ARIMA(my_fourth_root(wili) ~ pdq(2, 1, 0)))

sims <- fit_joint |>
  generate(h = 4, times = 1000, bootstrap = TRUE) |>
  group_by(location, .rep) |>
  mutate(horizon = row_number()) |>
  ungroup() |>
  as_tibble()
```

We reshape the samples for our target horizon into a matrix with one row per region and one column per sample, which is the format the scoring functions expect.

```{r partA-matrix}
samp_wide <- sims |>
  filter(horizon == !!horizon) |>
  select(location, .rep, .sim) |>
  pivot_wider(names_from = .rep, values_from = .sim)

target_end_date <- sims |>
  filter(horizon == !!horizon) |> pull(origin_date) |> unique()

dat_indep <- as.matrix(samp_wide[, -1])
rownames(dat_indep) <- samp_wide$location

## observed value at the target date, one per region
y <- flu_data_hhs |> as_tibble() |>
  filter(location %in% sel_locations, origin_date == target_end_date) |>
  arrange(match(location, samp_wide$location)) |>
  pull(wili)
names(y) <- samp_wide$location

target_end_date
y
```

# The energy score

The energy score is a proper scoring rule for a **vector** outcome $\mathbf{y}$, and it generalises the CRPS.
Writing $\mathbf{X}, \mathbf{X}'$ for independent draws from the forecast and $\lVert\cdot\rVert$ for Euclidean distance, it balances accuracy against sharpness exactly as the CRPS does, but reads the whole vector at once.
In one dimension it *is* the CRPS.

::: {.callout-tip .optional collapse="true"}
## Mathematical definition (optional)

$$
\text{ES}(F, \mathbf{y}) = \mathbb{E}\lVert \mathbf{X} - \mathbf{y} \rVert - \tfrac{1}{2}\,\mathbb{E}\lVert \mathbf{X} - \mathbf{X}' \rVert
$$

The first term rewards forecasts whose draws land close to the observed vector.
The second term rewards a concentrated cloud of draws.
To compute it we replace the expectations with averages over our predictive samples.
:::

::: callout-tip
### Key things to note about the energy score

-   Small values are better, and it is reported in the units of the data.
-   It is a genuine generalisation of the CRPS, to which it reduces in one dimension.
-   It needs **joint samples**, so a forecast that treats regions independently and one that couples them are scored differently.
-   Like the CRPS it is best used to **compare** forecasts, not read in isolation.
:::

::: {.callout-tip .optional collapse="true"}
## Interactive: the energy score

```{=html}
<iframe class="nfidd-embed" src="../interactive/multivariate-forecast-evaluation.html?embed=energy" title="Energy score interactive" loading="lazy" style="width:100%;height:820px;border:1px solid #e5e7eb;border-radius:10px;background:#fff"></iframe>
<script>(function(){if(window.__nfiddEmbedInit)return;window.__nfiddEmbedInit=1;window.addEventListener('message',function(ev){var h=ev&&ev.data&&ev.data.nfiddEmbedHeight;if(!h)return;document.querySelectorAll('iframe.nfidd-embed').forEach(function(f){if(f.contentWindow===ev.source)f.style.height=(h+24)+'px';});});})();</script>
```

Drag the observation and rotate the forecast cloud with the dependence slider.
The mean marginal CRPS never moves; only the energy score responds to the dependence.
:::

To see the difference dependence makes, we build a second forecast with **identical marginals** but perfect positive dependence.
Sorting each region's samples makes high draws in one region line up with high draws in the others.

```{r partA-dependence}
## identical marginals, but now the regions move together (comonotonic)
dat_comono <- t(apply(dat_indep, 1, sort))

## marginal CRPS (per region) is unchanged by the reordering
crps_indep  <- scoringRules::crps_sample(y = y, dat = dat_indep)
crps_comono <- scoringRules::crps_sample(y = y, dat = dat_comono)

rbind(independent = crps_indep, comonotonic = crps_comono)
c(mean_indep = mean(crps_indep), mean_comono = mean(crps_comono))
```

The per-region CRPS, and therefore the average, are exactly the same for both forecasts.
The energy score is not.

```{r partA-energy}
es_indep  <- scoringRules::es_sample(y = y, dat = dat_indep)
es_comono <- scoringRules::es_sample(y = y, dat = dat_comono)

c(independent = es_indep, comonotonic = es_comono)
```

::: callout-tip
## Take 5 minutes

The mean marginal CRPS is identical for the two forecasts, but the energy score prefers the comonotonic one.
Look at the observed vector `y` relative to the forecast medians.
Why does coupling the regions help here, and when would it hurt instead?
:::

::: {.callout-note collapse="true"}
## Solution

At this target date all three regions came in **below** their forecast medians at the same time, a synchronised late-season decline.
The comonotonic forecast puts probability on regions being jointly low (or jointly high), so it places mass near what actually happened and scores better.
The independent forecast, which is what fitting each region separately gives us, spreads its joint mass across combinations where one region is high while another is low, and scores worse for it.

If instead the regions had diverged, one high while another was low, the coupling would have hurt and the independent forecast would have scored better.
The marginal CRPS is blind to all of this: it is identical either way.
The lesson is not that dependence is always good, but that the energy score can see it and marginal scores cannot.
:::

# Going further

## Challenge

-   Fit a genuinely joint-across-locations model with `fable::VAR()` and compare its energy score to an independent ARIMA forecast.
-   Add the variogram score (another multivariate score, available via `scoringRules::vs_sample()`) to the joint comparison. How does it differ from the energy score in what it rewards?

## Methods in practice

-   The `scoringRules` package [documentation](https://cran.r-project.org/package=scoringRules) covers multivariate scores including the energy and variogram scores.
-   Hubs increasingly collect sample (trajectory) forecasts rather than only quantiles, which keeps the joint distribution across locations and horizons.[@hubverseModelOutput2026]
-   @bay_evaluation_2026 describe using the energy score to assess trajectories in longer-term scenario predictions.
-   @macarthur_collaborative_2026 use an adapted energy score to evaluate nowcasts of the proportions of SARS-CoV-2 variants circulating at different timepoints.

# Wrap up

-   Review what you've learned in this session with the [learning objectives](../reference/learning_objectives.qmd)

::: {#refs}
:::
