---
title: "Baselines and seasonality"
author: "Nowcasting and forecasting of infectious disease dynamics"
engine: knitr
format:
  revealjs:
    output: slides/baselines-and-seasonality.html
    footer: "Baselines and seasonality"
    chalkboard: true
    slide-level: 3
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library("nfidd.forecasting")
library("fable")
library("dplyr")
library("ggplot2")
library("ggtime")
theme_set(theme_bw())

data(flu_data)
flu_data <- flu_data |>
  filter(epiweek <= "2018-01-01") # middle of the 2017/2018 season

fourth_root <- function(x) x^0.25
inv_fourth_root <- function(x) x^4
my_fourth_root <- new_transformation(fourth_root, inv_fourth_root)

simple_fits <- flu_data |>
  model(
    persistence = NAIVE(my_fourth_root(wili)),
    mean        = MEAN(my_fourth_root(wili))
  )
fit_fourier <- flu_data |>
  model(fourier = ARIMA(my_fourth_root(wili) ~ pdq(0, 0, 0) + fourier(period = "year", K = 3)))
fit_ar2f <- flu_data |>
  model(arima200_fourier = ARIMA(my_fourth_root(wili) ~ pdq(2, 0, 0) + fourier(period = "year", K = 3)))
all_fits <- flu_data |>
  model(
    persistence      = NAIVE(my_fourth_root(wili)),
    mean             = MEAN(my_fourth_root(wili)),
    fourier          = ARIMA(my_fourth_root(wili) ~ pdq(0, 0, 0) + fourier(period = "year", K = 3)),
    arima200_fourier = ARIMA(my_fourth_root(wili) ~ pdq(2, 0, 0) + fourier(period = "year", K = 3))
  )

recent <- flu_data |> filter(epiweek >= as.Date("2015-09-01"))
```

### Good — compared to *what*?

Before asking "is my model any good?", ask **"good compared to what?"**

A **baseline** (reference) model is a deliberately simple, naive forecast that a useful model ought to beat.

- Sets a floor — beat it, or something's wrong
- Simple assumptions, computationally cheap, hard to overfit
- A *shared* baseline lets everyone measure "skill" on the same yardstick

### Three simple baselines

- **Persistence** — "nothing changes", predicts the most recent observation
- **Mean** — "reverts to normal", predicts the long-run average
- **Seasonal** — "this time of year looks like it usually does", predicts the typical seasonal pattern

We keep the fourth-root transformation so every model is on the same footing.

### Persistence & mean

```{r fc-simple, echo=FALSE, fig.height=3.2}
forecast(simple_fits, h = 40) |>
  autoplot(recent) +
  facet_wrap(~.model) +
  labs(y = "% of visits")
```

Persistence extends the last value; the mean ignores where we are. 

### The data are seasonal

```{r season-plot, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3.2}
flu_data |>
  gg_season(wili, period = "year", labels = "none") +
  ggtitle("Every season has a similar winter-peaked shape")
```

### A seasonal baseline: harmonic regression

`ARIMA(y ~ pdq(0,0,0) + fourier(period = "year", K = 3))`

- **No** AR / differencing / MA — *only* sine & cosine curves with a 1-year period
- Forecast = a smooth "seasonal climatology"; still ignores the most recent weeks

```{r fc-fourier, echo=FALSE, fig.height=2.8}
forecast(fit_fourier, h = 40) |>
  autoplot(recent) +
  labs(y = "% of visits")
```

### Improving on the baselines

`ARIMA(y ~ pdq(2,0,0) + fourier(period = "year", K = 3))`

- AR(2) anchors to recent values; Fourier pulls toward the seasonal shape

```{r fc-ar2f, echo=FALSE, fig.height=2.8}
forecast(fit_ar2f, h = 40) |>
  autoplot(recent) +
  labs(y = "% of visits")
```

### All together

```{r fc-all, echo=FALSE, fig.height=3.4}
forecast(all_fits, h = 40) |>
  autoplot(recent) +
  facet_wrap(~.model) +
  labs(y = "% of visits")
```

The combined model captures the seasonal shape *and* starts from the right level — but "looks better" isn't proof. 

### Did it work? Check the residuals

If a model captured the structure, its residuals should look like **white noise** — flat ACF, especially at the seasonal lag ~52.

```{r resid, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3.2}
fit_ar2f |> gg_tsresiduals(lag_max=55)
```

::: {.callout-note title="What we want to see"}

 - top panel: white noise in the residuals
 - bottom left: no significant residual autocorrelation
 - bottom right: Normal, symmetric distrbution of residuals

:::

## `r fontawesome::fa("laptop-code", "white")` Your Turn {background-color="#447099" transition="fade-in"}

1.  Build the three baselines: `RW()`, `MEAN()`, and an `fourier()` seasonal model.
2.  Add recent dynamics with an `ARIMA() + fourier()` model and compare against the baselines.
3.  Use `gg_tsresiduals()` to check whether the model captured the structure.

#

[Return to the session](../baselines-and-seasonality.qmd)
