---
title: "Selecting test-phase models"
author: "Nowcasting and forecasting of infectious disease dynamics"
engine: knitr
format:
  revealjs:
    output: slides/hub-playground-testing.html
    footer: "Selecting test-phase models"
    chalkboard: true
    slide-level: 3
bibliography: ../../nfidd.bib
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library("nfidd.forecasting")
library("dplyr")
library("ggplot2")
theme_set(theme_bw(base_size = 15))
```

```{=html}
<style>
.reveal .mermaid { display: flex; justify-content: center; }
.reveal .mermaid svg { width: 100%; max-width: 1100px !important; height: auto; }
</style>
```

### From validation to testing

- You built **many** models on the **two validation seasons**.
- Now: choose just **1–2** to carry into the **three testing seasons**.
- The testing seasons stand in for **real-time** forecasting.

```{r split-fig, echo=FALSE, fig.width=9, fig.height=3}
data(flu_data_hhs)
nat <- flu_data_hhs |>
  tibble::as_tibble() |>
  filter(location == "US National") |>
  mutate(origin_date = as.Date(origin_date))

seasons <- tibble::tribble(
  ~season,     ~start,       ~end,         ~phase,
  "2015/2016", "2015-10-01", "2016-05-15", "Validation",
  "2016/2017", "2016-10-01", "2017-05-15", "Validation",
  "2017/2018", "2017-10-01", "2018-05-15", "Testing",
  "2018/2019", "2018-10-01", "2019-05-15", "Testing",
  "2019/2020", "2019-10-01", "2020-05-15", "Testing"
) |>
  mutate(across(c(start, end), as.Date),
         phase = factor(phase, levels = c("Validation", "Testing")))

xlim <- as.Date(c("2015-06-01", "2020-08-01"))
ggplot() +
  geom_rect(data = seasons,
            aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf, fill = phase),
            alpha = 0.25) +
  geom_line(data = filter(nat, origin_date >= xlim[1], origin_date <= as.Date("2017-07-01")),
            aes(origin_date, wili), linewidth = 0.5) +
  scale_fill_manual(values = c(Validation = "#1b9e77", Testing = "#d95f02"), name = NULL) +
  scale_x_date(limits = xlim, date_breaks = "1 year", date_labels = "%Y") +
  labs(x = NULL, y = "ILI %") +
  theme(legend.position = "top")
```

### The temptation

From the earlier session: you have several (or many) fitted models and their validation scores.

Tempting move: **submit the one with the best validation score.**

::: {.fragment .fade-in}
::: callout-caution

But — will it still be the best on seasons it has *never seen*?
:::
:::

### Validation rank ≠ test rank

- Each forecast is genuinely **out-of-sample in time** — the model never sees the weeks it predicts.
- **But** across the validation seasons you iterate: tweak, refit, re-score, keep what looks best.
- So those seasons quietly become **in-sample for model *selection*** — you are, in effect, fitting to them.
- And it's a **tiny sample — just two seasons** — so the rankings are **noisy** and **optimistically biased**.

### If you've seen one flu season...

The saying at CDC goes:

> If you've seen one flu season... you've seen one flu season.

In other words, every season is unique!

::: callout-caution
This further suggests that we need to be cautious about assuming that validation phase patterns will be the same in the unpeeked-at testing seasons.
:::

### How to choose {.smaller}

Favour models that are:

- ✅ **consistently good** in CV — not just the single lowest score
- ✅ **parsimonious** — fewer parameters overfit less
- ✅ **diverse** — e.g. best `ARIMA`, best `NNETAR`, best ensemble
- ✅ **stable** across seasons *and* locations

Then pass only **1–2** to the testing phase.

::: callout-tip

The best-*looking* model isn't reliably the best *forecaster*.
:::


### Your selection rule is a model too

- Every peek at validation to pick a model is a kind of **fitting**.
- Decide your rule, then **commit** — don't tune on the testing seasons.
- A **simpler** selection process → a more trustworthy test result.

### The testing phase = real-time

- Treat 2017/18–2019/20 as if it were **live**:
  **generate once, no refitting, no peeking.**

::: {.fragment .fade-in}
::: callout-note

Our cross-validation uses the **vintage** data available at each date, so — like the original hub models — you forecast from real-time data, not finalized data.
:::
:::

### Why submit to an online hub?

- A shared **format** → automatic validation of your submission
- Line your forecasts up against **others** — and build **ensembles**
- The reward: your model on the **[public dashboard](https://reichlab.io/sismid-ili-forecasting-dashboard/)**

### Two ways to submit

::: {.fragment .fade-in}
**Get help** — hand your files to an instructor/TA/classmate to upload
:::

::: {.fragment .fade-in}
**Drag & drop** — Upload files using web interface to GitHub.com.
*(you still need a GitHub account)*
:::

::: {.fragment .fade-in}
**Fork → clone → PR** — the full GitHub workflow, on your own
:::

### The pull-request loop (if using GitHub)

```{mermaid}
%%{init: {'theme':'base', 'themeVariables': {'fontSize':'26px'}, 'flowchart': {'curve':'basis', 'nodeSpacing':45, 'rankSpacing':55}}}%%
flowchart LR
  A["Fork the hub"] --> B["Clone your fork"]
  B --> C["Add forecasts<br/>+ metadata"]
  C --> D["Commit & push"]
  D --> E["Open a PR"]
  E --> F{"Checks<br/>pass?"}
  F -- no --> C
  F -- yes --> G["Merged →<br/>dashboard 🎉"]
```

Watch the automated checks; fix any errors they flag.

## `r fontawesome::fa("laptop-code", "white")` Your Turn {background-color="#447099" transition="fade-in"}

1. **Select** 1–2 models from your validation work.
2. **Generate** test-phase forecasts (seasons 2017/18–2019/20).
3. **Submit** them to the hub via a pull request.
4. **Find yourself** on the dashboard!

#

[Return to the session](../hub-playground-testing.qmd)

### References

::: {#refs}
:::
